---
title: "Environmental data sources"
---

## Overview

This chapter documents the data sources used to construct the environmental and geographic predictors in the EPP climate analysis. It separates two things: the source products from which variables were derived, and the repository files needed to reproduce the analyses from the processed data.

## Data sources

```{r}
#| label: main-environmental-data-sources
#| echo: true

source_table <- data.frame(
  `Data component` = c(
    "EPP records and metadata",
    "Monthly climate",
    "Climate anomalies",
    "NDVI",
    "Distance to coast",
    "Taxonomy",
    "Phylogeny"
  ),
  `Variables or objects used` = c(
    "n_epp_broods, n_broods_sampled, species, population, location, breeding timing",
    "tmp_*, pre_*, dtr_*",
    "short- and long-term temperature and precipitation anomaly variables",
    "ndvi_first_month_median_30km; ndvi_full_season_median_30km",
    "distance_to_coast_km",
    "Clements_name; species_phylo; species_nonphylo",
    "summary_dated_clements.nex; phylo_mat"
  ),
  `Source or extraction route` = c(
    "Compiled EPP dataset prepared for this project",
    "CRU monthly climate products for the breeding-season month or season",
    "Derived from CRU climate products relative to local baseline conditions",
    "NOAA NDVI extracted around study coordinates in Google Earth Engine",
    "Derived from study coordinates",
    "Clements Checklist v2025",
    "Clements-matched dated avian phylogeny"
  ),
  `Repository status` = c(
    "Included in final analysis-ready files",
    "Derived variables included; raw CRU files not redistributed",
    "Derived variables included; raw CRU files not redistributed",
    "Derived variables included; Google Earth Engine products not redistributed",
    "Final values included",
    "Matched taxonomy included",
    "Tree/covariance objects included where needed for modelling"
  ),
  check.names = FALSE
)

knitr::kable(source_table)
```

## Files needed to reproduce the analysis

```{r}
#| label: reproducibility-files
#| echo: true

reproducibility_files <- data.frame(
  `File or folder` = c(
    "epp_data_June2026.csv",
    "metadata_June2026.csv",
    "data/processed/dat_model_phylo.rds",
    "data/processed/phylo_mat.rds",
    "phylogeny/summary_dated_clements.nex",
    "models/",
    "figures/",
    "*.qmd files and _quarto.yml",
    "scripts/"
  ),
  `Purpose` = c(
    "Final analysis-ready dataset containing EPP records, environmental predictors, geography, taxonomy, and derived covariates",
    "Variable definitions and dataset metadata",
    "Model-ready dataset with phylogenetic species identifiers",
    "Phylogenetic covariance matrix used in Bayesian models",
    "Source phylogenetic tree used to construct the covariance matrix",
    "Saved model summaries, diagnostics, LOO outputs, and lightweight model outputs used by the book",
    "Saved figures called by the Quarto chapters",
    "Book chapters and rendering configuration",
    "R scripts used to fit models and create saved outputs"
  ),
  check.names = FALSE
)

knitr::kable(reproducibility_files)
```

The analyses reported in this project can be reproduced from the processed files included in the repository. Full re-extraction of environmental variables from raw source products would additionally require access to the original CRU climate files and the Google Earth Engine products used for NDVI extraction.

## Computational environment

```{r}
#| label: computational-environment
#| echo: true

environment_summary <- data.frame(
  item = c(
    "R version",
    "Platform",
    "Operating system"
  ),
  value = c(
    R.version.string,
    R.version$platform,
    Sys.info()[["sysname"]]
  )
)

knitr::kable(environment_summary)
```

