---
title: "Environmental data sources"
---

## Overview

:::: {.callout-note title="Repository Principle"}
::: {.callout-note title="Repository Principle"}
This chapter documents the environmental and geographic data sources used in the analysis of extra-pair paternity (EPP) in birds. The repository contains the final analysis-ready dataset, metadata, model code, and workflow documentation. Large external climate and remote-sensing products are documented here, but are not redistributed.

The analyses reported in this project can be reproduced from the processed files included in the repository. Full re-extraction of the environmental variables from raw source products would additionally require access to the original CRU climate files and to the Google Earth Engine products used for NDVI extraction.
:::
::::

The final dataset used in the analyses is:

``` text
epp_data_June2026.csv
```

Variable definitions are provided in:

``` text
metadata_June2026.csv
```

These files contain the EPP records, taxonomy, geographic information, breeding-season timing, environmental variables, and derived covariates used by the models.

## Main environmental data sources

```{r}
#| label: main-environmental-data-sources
#| echo: false

source_table <- data.frame(
  `Data component` = c(
    "Monthly climate",
    "NDVI",
    "Distance to coast",
    "Taxonomy",
    "Phylogeny"
  ),
  `Variables in the final dataset` = c(
    "tmp_*, pre_*, dtr_*, short- and long-term anomaly variables",
    "ndvi_first_month_median_30km; ndvi_full_season_median_30km",
    "distance_to_coast_km",
    "Clements_name",
    "species_phylo; phylo_mat"
  ),
  `Source and extraction route` = c(
    "CRU monthly climate for the breeding-season month or season",
    "NOAA NDVI extracted around study coordinates in Google Earth Engine",
    "Derived from study coordinates",
    "Clements Checklist v2025",
    "Clements-matched tree: phylogeny/summary_dated_clements.nex"
  ),
  `Repository status` = c(
    "Derived variables included; raw CRU files not redistributed",
    "Derived variables included; extraction script provided",
    "Final values included",
    "Matched taxonomy included",
    "Tree file included; phylo_mat saved in data/processed/phylo_mat.rds"
  ),
  check.names = FALSE
)

knitr::kable(source_table)
```

## CRU climate variables

Monthly climate variables were derived from external CRU climate data before being merged into the final EPP dataset. The raw CRU files are not included in the GitHub repository because they are external data products and may be large, versioned, or subject to their own distribution terms.

The CRU data source is documented here:

``` text
https://crudata.uea.ac.uk/cru/data/hrg/
```

The extraction used the CRU folders/products corresponding to:

- `tmp` - mean temperature;
- `pre` - precipitation;
- `dtr` - diurnal temperature range.

The final dataset includes the derived climate variables needed for the analyses:

- mean temperature for the selected breeding-season window and the full breeding season;
- precipitation for the selected breeding-season window and the full breeding season;
- diurnal temperature range for the selected breeding-season window and the full breeding season;
- short-term 5-year temperature and precipitation anomalies;
- long-term 1961-1989 temperature and precipitation anomalies.

::: {.callout-important title="Climate Reproducibility"}
The published analyses can be rerun from `epp_data_June2026.csv` without downloading the raw CRU files. Full re-extraction of the climate variables would require access to the same external CRU data and the local extraction workflow used to generate the derived variables.
:::

## NDVI extraction

NDVI was used as a proxy for habitat productivity around each study location. The final variables describe median NDVI within a 30 km buffer around the EPP study coordinates for the selected breeding-season window and, where relevant, the full breeding season.

The NDVI values were extracted using Google Earth Engine. The JavaScript script used for the NDVI extraction is provided in the GitHub repository with the environmental extraction scripts.

The final NDVI variables included in `epp_data_June2026.csv` are:

- `ndvi_first_month_median_30km`;
- `ndvi_full_season_median_30km`;

## Taxonomy and phylogeny

Species names were matched using the Clements Checklist v2025:

``` text
data_raw/Clements_v2025-October-2025.csv
```

Species names were then linked to the Clements-matched phylogenetic tree described by McTavish et al. (2025):

``` text
phylogeny/summary_dated_clements.nex
```

The tree was pruned to the retained species and converted to `data/processed/phylo_mat.rds`. The phylogenetic effect in the Bayesian models is:

``` text
(1 | gr(species_phylo, cov = phylo_mat))
```

Source citation:

``` text
McTavish, E. J., Gerbracht, J. A., Holder, M. T., Iliff, M. J., Lepage, D., Rasmussen, P. C., ... & Miller, E. T. (2025). A complete and dynamic tree of birds. Proceedings of the National Academy of Sciences, 122(18), e2409658122.
```

## Reproducibility

To reproduce the models reported in this project, a user needs:

- `epp_data_June2026.csv`;
- `metadata_June2026.csv`;
- `phylogeny/summary_dated_clements.nex`
- `data/processed/dat_model_phylo.rds`;
- `data/processed/phylo_mat.rds`;
- the Quarto files documenting the workflow.

To reproduce the environmental extraction from raw source products, a user would additionally need access to the CRU climate files and the cloud-hosted Google Earth Engine products used for NDVI.

## Computational Environment

The analyses and Quarto book were prepared in R. The exact R version and platform used for rendering are reported below.

```{r}
#| label: computational-environment
#| echo: false

environment_summary <- data.frame(
  item = c(
    "R version",
    "Platform",
    "Operating system"
  ),
  value = c(
    R.version.string,
    R.version$platform,
    Sys.info()[["sysname"]]
  )
)

knitr::kable(environment_summary)
```

The main R packages used across the workflow include `brms`, `dplyr`, `ggplot2`, `posterior`, `ape`, `phytools`, `knitr`, and `quarto`.
