---
title: "Taxonomy and phylogeny"
---

# Aim

This page prepares the taxonomic and phylogenetic structure used in the Bayesian comparative analyses.

The goal is to verify species matching between the analytical dataset and the phylogenetic tree, prune the tree to the species included in the study, generate a phylogenetic correlation matrix, and prepare the phylogenetic grouping variables required for the Bayesian models.

# Load packages and data

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(ape)
library(knitr)

dat_model <- readRDS("data/processed/dat_model.rds")

dat_taxonomy <- readr::read_csv(
  "epp_data_June2026.csv",
  show_col_types = FALSE
) %>%
  select(
    any_of(
      c(
        "epp_row_id",
        "record_id",
        "scientific_name",
        "Clements_name",
        "order",
        "family",
        "family_common"
      )
    )
  ) %>%
  distinct()

if (
  !"family" %in% names(dat_model) ||
    ("family" %in% names(dat_model) && all(is.na(dat_model$family)))
) {
  dat_model <- dat_model %>%
    select(-any_of(c("order", "family", "family_common")))
}

needs_family_merge <- !"family" %in% names(dat_model) ||
  all(is.na(dat_model$family))

if (needs_family_merge && "epp_row_id" %in% names(dat_model)) {
  dat_model <- dat_model %>%
    left_join(
      dat_taxonomy %>%
        select(
          epp_row_id,
          any_of(c("order", "family", "family_common"))
        ),
      by = "epp_row_id"
    )
}

needs_family_merge <- !"family" %in% names(dat_model) ||
  all(is.na(dat_model$family))

if (needs_family_merge && "record_id" %in% names(dat_model)) {
  dat_model <- dat_model %>%
    left_join(
      dat_taxonomy %>%
        select(
          record_id,
          any_of(c("order", "family", "family_common"))
        ),
      by = "record_id"
    )
}

needs_family_merge <- !"family" %in% names(dat_model) ||
  all(is.na(dat_model$family))

if (needs_family_merge && "Clements_name" %in% names(dat_model)) {
  dat_model <- dat_model %>%
    left_join(
      dat_taxonomy %>%
        select(
          Clements_name,
          any_of(c("order", "family", "family_common"))
        ) %>%
        distinct(),
      by = "Clements_name"
    )
}

tree <- ape::read.nexus("phylogeny/summary_dated_clements.nex")
```

# Phylogenetic tree overview

The phylogenetic analyses are based on the global avian phylogeny using Clements taxonomy.

```{r}
#| label: tree-overview

tree
```

# Species included in the analytical dataset

The cleaned EPP dataset contains species names matched to the Clements taxonomic backbone. These names are stored in the variable `Clements_name`.

```{r}
#| label: species-summary

species_data <- dat_model %>%
  distinct(Clements_name) %>%
  filter(!is.na(Clements_name)) %>%
  pull(Clements_name)

species_tree <- tree$tip.label

species_summary <- tibble(
  metric = c(
    "Species in analytical dataset",
    "Tips in full phylogeny"
  ),
  value = c(
    length(species_data),
    length(species_tree)
  )
)

knitr::kable(species_summary)
```

# Check taxonomic matching

We verified that all species included in the analytical dataset were represented in the phylogenetic tree.

```{r}
#| label: species-names-check

missing_from_tree <- setdiff(
  species_data,
  species_tree
)

missing_from_tree
```

```{r}
#| label: missing-summary

tibble(
  n_species_missing_from_tree = length(missing_from_tree)
) %>%
  knitr::kable()
```

A value of zero indicates complete taxonomic matching between the dataset and the phylogeny.

# Prune the phylogeny

The full phylogenetic tree contains many species not included in the EPP dataset. Therefore, the tree was pruned to retain only species present in the analytical dataset.

```{r}
#| label: prune-tree

tree_pruned <- ape::drop.tip(
  tree,
  setdiff(tree$tip.label, species_data)
)
```

```{r}
#| label: pruned-tree-overview

tree_pruned
```

```{r}
#| label: pruned-tree-summary

tibble(
  metric = c(
    "Species retained in analytical dataset",
    "Tips in pruned phylogeny"
  ),
  value = c(
    length(species_data),
    length(tree_pruned$tip.label)
  )
) %>%
  knitr::kable()
```

# Generate the phylogenetic correlation matrix

The Bayesian models account for phylogenetic non-independence among species through a phylogenetic random effect.

To implement this structure in `brms`, the pruned phylogeny was converted into a phylogenetic correlation matrix.

```{r}
#| label: phylogenetic-correlation-matrix

phylo_mat <- ape::vcv.phylo(
  tree_pruned,
  corr = TRUE
)

dim(phylo_mat)
```

```{r}
#| label: phylo-matrix-preview

phylo_mat[1:5, 1:5]
```

# Prepare phylogenetic grouping variables

Two species-level grouping variables were retained for the Bayesian models:

- `species_phylo` - species identity linked to the phylogenetic covariance matrix;
- `species_nonphylo` - species identity modelled as an additional non-phylogenetic random effect.

Including both random effects allows the models to partition variation attributable to shared evolutionary history and species-specific variation not explained by phylogeny.

```{r}
#| label: add-phylo-columns

dat_model_phylo <- dat_model %>%
  mutate(
    species_phylo = Clements_name,
    species_nonphylo = scientific_name
  ) %>%
  filter(
    species_phylo %in% tree_pruned$tip.label
  )
```

# Final Dataset Summary

```{r}
#| label: phylo-dataset-summary

dat_model_phylo %>%
  summarise(
    n_records = n(),
    n_studies = n_distinct(id),
    n_species_phylo = n_distinct(species_phylo),
    n_species_nonphylo = n_distinct(species_nonphylo),
    n_families = n_distinct(family),
    n_orders = n_distinct(order),
    n_populations = n_distinct(population_id)
  ) %>%
  knitr::kable()
```

# Save Phylogenetic Objects

The processed phylogenetic objects are saved for subsequent modelling steps.

```{r}
#| label: save-phylo-objects

saveRDS(
  dat_model_phylo,
  "data/processed/dat_model_phylo.rds"
)

saveRDS(
  tree_pruned,
  "data/processed/tree_pruned.rds"
)

saveRDS(
  phylo_mat,
  "data/processed/phylo_mat.rds"
)
```

# Output Files

| File | Description |
|------------------------------------|------------------------------------|
| `dat_model_phylo.rds` | Analytical dataset with phylogenetic grouping variables |
| `tree_pruned.rds` | Pruned phylogenetic tree containing only species included in the study |
| `phylo_mat.rds` | Phylogenetic correlation matrix used in Bayesian models |
