---
title: "Overall model structure"
---

## Overview

This chapter defines the common statistical structure used across the Bayesian models. It describes the beta-binomial response model, the population, species, and phylogenetic grouping terms used for the expected EPP probability, and the distinction between mean-only and location-scale models. Reference identity (`id_reference`) is not included as a separate grouping term in the model because the reference-population overlap check in the data-preparation chapter showed that population identity captures the repeated-observation structure used in the main environmental models.

```{r}
#| label: model-structure-setup
#| echo: true
#| message: false
#| warning: false

library(dplyr)
library(readr)
library(tibble)
library(knitr)

first_existing <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) {
    return(NA_character_)
  }
  existing[1]
}

overall_summary_file <- first_existing(c(
  "models/overall_population/summary_m_overall_population.txt",
  "models/overall_model/summary_m_overall_population.txt",
  "models/summary_m_overall_population.txt"
))

overall_diagnostics_file <- first_existing(c(
  "models/overall_population/overall_population_model_diagnostics.csv",
  "models/overall_population/diagnostics_m_overall_population.csv",
  "models/overall_model/overall_model_diagnostics.csv"
))

overall_model_file <- first_existing(c(
  "models/overall_population/m_overall_population.rds",
  "models/overall_model/m_overall_population.rds",
  "models/m_overall_population.rds"
))
```

## Response model

For record $i$, the response is the number of sampled broods with EPP:

$$y_i = \texttt{n\_epp\_broods}_i,$$

out of the total number of sampled broods:

$$n_i = \texttt{n\_broods\_sampled}_i.$$

The model likelihood is:

$$y_i \sim \operatorname{BetaBinomial}(n_i, \mu_i, \phi_i),$$

where $\mu_i$ is the expected EPP probability and $\phi_i$ is the beta-binomial precision parameter.

## Expected EPP and grouping structure

The baseline component for the expected EPP probability is:

$$\operatorname{logit}(\mu_i) =
\alpha +
\mathbf{x}_i \boldsymbol{\beta} +
u_{\operatorname{population}[i]} +
v_{\operatorname{species}[i]} +
w_{\operatorname{phylogeny}[i]}.$$

The grouping terms correspond to population identity, non-phylogenetic species identity, and phylogenetic species covariance:

$$u_{\operatorname{population}[i]} \sim \mathcal{N}(0, \sigma^2_{\operatorname{population}}),$$

$$v_{\operatorname{species}[i]} \sim \mathcal{N}(0, \sigma^2_{\operatorname{species}}),$$

$$w_{\operatorname{phylogeny}} \sim \mathcal{N}(0, \sigma^2_{\operatorname{phylogeny}}\mathbf{A}),$$

where $\mathbf{A}$ is the phylogenetic correlation matrix.

Here $\mathbf{x}_i \boldsymbol{\beta}$ denotes the fixed-effect part of the model. The exact predictors differ between the environmental models, but the linear predictor always combines fixed effects with the same random-effect structure: `population_id`, `species_nonphylo`, and `species_phylo` with the phylogenetic covariance matrix.

In `brms` notation, this is:

``` text
(1 | population_id) +
(1 | species_nonphylo) +
(1 | gr(species_phylo, cov = phylo_mat))
```

## Precision parameter

The parameter $\phi_i$ is the beta-binomial precision parameter. It controls the amount of residual variation around the expected EPP probability $\mu_i$. Lower precision corresponds to stronger extra-binomial variation among records with the same expected EPP probability. Positive coefficients in the precision component indicate higher precision and lower residual overdispersion; negative coefficients indicate lower precision and higher residual overdispersion.

In mean-only models, the precision parameter is constant across records:

$$\log(\phi_i) = \gamma_0.$$

In mean-scale models, selected predictors are also included in the precision component:

$$\log(\phi_i) =
\gamma_0 +
\mathbf{z}_i \boldsymbol{\gamma}.$$

Here $\mathbf{z}_i \boldsymbol{\gamma}$ denotes the fixed-effect part of the scale model. Thus, mean-only models ask whether predictors explain average EPP, whereas mean-scale models additionally ask whether predictors explain variation in the precision of EPP estimates.

## Baseline overall model

The baseline overall model uses only the common grouping structure, without environmental predictors:

$$\operatorname{logit}(\mu_i) =
\alpha +
u_{\operatorname{population}[i]} +
v_{\operatorname{species}[i]} +
w_{\operatorname{phylogeny}[i]}.$$

The precision component is:

$$\log(\phi_i) = \gamma_0.$$

In `brms` notation:

```{r}
#| label: overall-model-formula
#| eval: false

m_overall_population <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    1 +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_model,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log")
)
```

## Overall model summary

```{r}
#| label: overall-model-result
#| echo: true
#| results: asis

if (!is.na(overall_summary_file)) {
  cat("<details><summary>Show overall model summary</summary>\n\n```\n")
  cat(readLines(overall_summary_file), sep = "\n")
  cat("\n```\n</details>\n")
} else {
  invisible(NULL)
}
```

## Overall model diagnostics

```{r}
#| label: overall-model-diagnostics
#| echo: true

if (!is.na(overall_diagnostics_file)) {
  read_csv(overall_diagnostics_file, show_col_types = FALSE) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
    knitr::kable()
} else {
  invisible(NULL)
}
```

