---
title: "Overall model structure"
---

## Overview

This chapter defines the common statistical structure used across the Bayesian models. It describes the beta-binomial response model, the population, species, and phylogenetic grouping terms used for the expected EPP probability, and the distinction between mean-only and location-scale models. Reference identity (`id_reference`) is not included as a separate grouping term: in the 571-record response-complete dataset, 223 of 230 populations (96.96%) were represented by a single publication. Population identity therefore captures nearly all of the repeated-observation structure available on the population side, while including both study- and population-level random intercepts would make their variances weakly identifiable. This omission of the preregistered study effect and its empirical justification are reported explicitly in the data-preparation chapter.

```{r}
#| label: model-structure-setup
#| echo: true
#| message: false
#| warning: false

library(dplyr)
library(readr)
library(tibble)
library(knitr)

read_csv_safe <- function(path) {
  tibble::as_tibble(utils::read.csv(
    path, check.names = FALSE, stringsAsFactors = FALSE
  ))
}

first_existing <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) {
    return(NA_character_)
  }
  existing[1]
}

overall_summary_file <- first_existing(c(
  "models/overall_population/summary_m_overall_population.txt",
  "models/overall_model/summary_m_overall_population.txt",
  "models/summary_m_overall_population.txt"
))

overall_diagnostics_file <- first_existing(c(
  "models/overall_population/overall_population_model_diagnostics.csv",
  "models/overall_population/diagnostics_m_overall_population.csv",
  "models/overall_model/overall_model_diagnostics.csv"
))

overall_model_file <- first_existing(c(
  "models/overall_population/m_overall_population.rds",
  "models/overall_model/m_overall_population.rds",
  "models/m_overall_population.rds"
))

overall_variance_file <- first_existing(c(
  "models/overall_population/overall_population_variance_components.csv",
  "models/overall_model/overall_population_variance_components.csv"
))

overlap_summary <- read_csv_safe(
  "models/overall_population/reference_population_overlap_summary.csv"
)
overlap_values <- setNames(overlap_summary$value, overlap_summary$metric)
if (overlap_values[["Populations"]] != 230L ||
    overlap_values[["Populations linked to one reference"]] != 223L) {
  stop("Reference-population overlap no longer reproduces 223 of 230 populations.")
}

# Fail the render if the documented mean-only priors or sampler settings drift
# from the active primary-model script.
main_model_expressions <- parse("scripts/run_main_environmental_models.R")
is_assignment_to <- function(expression, object_name) {
  is.call(expression) &&
    identical(expression[[1]], as.name("<-")) &&
    identical(expression[[2]], as.name(object_name))
}
prior_assignment <- main_model_expressions[
  vapply(
    main_model_expressions,
    is_assignment_to,
    logical(1),
    object_name = "priors_mu"
  )
]
expected_priors_mu <- quote(c(
  prior(normal(0, 2), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "phi")
))
if (length(prior_assignment) != 1L ||
    !identical(prior_assignment[[1]][[3]], expected_priors_mu)) {
  stop("The documented priors_mu block does not match the active main-model script.")
}

main_model_lines <- trimws(readLines("scripts/run_main_environmental_models.R"))
required_sampler_lines <- c(
  "chains <- 4",
  "cores <- 4",
  "iter <- 10000",
  "warmup <- 5000",
  "control = list(adapt_delta = 0.99, max_treedepth = 15),"
)
missing_sampler_lines <- setdiff(required_sampler_lines, main_model_lines)
if (length(missing_sampler_lines)) {
  stop(
    "The documented sampler settings do not match the active main-model script: ",
    paste(missing_sampler_lines, collapse = "; ")
  )
}
```

## Response model

For record $i$, the response is the number of sampled broods with EPP:

$$y_i = \texttt{n\_epp\_broods}_i,$$

out of the total number of sampled broods:

$$n_i = \texttt{n\_broods\_sampled}_i.$$

The model likelihood is:

$$y_i \sim \operatorname{BetaBinomial}(n_i, \mu_i, \phi_i),$$

where $\mu_i$ is the expected EPP probability and $\phi_i$ is the beta-binomial precision parameter.

## Expected EPP and grouping structure

The baseline component for the expected EPP probability is:

$$\operatorname{logit}(\mu_i) =
\alpha +
\mathbf{x}_i \boldsymbol{\beta} +
u_{\operatorname{population}[i]} +
v_{\operatorname{species}[i]} +
w_{\operatorname{phylogeny}[i]}.$$

The grouping terms correspond to population identity, non-phylogenetic species identity, and phylogenetic species covariance:

$$u_{\operatorname{population}[i]} \sim \mathcal{N}(0, \sigma^2_{\operatorname{population}}),$$

$$v_{\operatorname{species}[i]} \sim \mathcal{N}(0, \sigma^2_{\operatorname{species}}),$$

$$w_{\operatorname{phylogeny}} \sim \mathcal{N}(0, \sigma^2_{\operatorname{phylogeny}}\mathbf{A}),$$

where $\mathbf{A}$ is the phylogenetic correlation matrix.

Here $\mathbf{x}_i \boldsymbol{\beta}$ denotes the fixed-effect part of the model. The exact predictors differ between the environmental models, but the linear predictor always combines fixed effects with the same random-effect structure: `population_id`, `species_nonphylo`, and `species_phylo` with the phylogenetic covariance matrix.

In `brms` notation, this is:

``` text
(1 | population_id) +
(1 | species_nonphylo) +
(1 | gr(species_phylo, cov = phylo_mat))
```

## Precision parameter

The parameter $\phi_i$ is the beta-binomial precision parameter. It controls the amount of residual variation around the expected EPP probability $\mu_i$. Lower precision corresponds to stronger extra-binomial variation among records with the same expected EPP probability. Positive coefficients in the precision component indicate higher precision and lower residual overdispersion; negative coefficients indicate lower precision and higher residual overdispersion.

In mean-only models, the precision parameter is constant across records:

$$\log(\phi_i) = \gamma_0.$$

In mean-scale models, selected predictors are also included in the precision component:

$$\log(\phi_i) =
\gamma_0 +
\mathbf{z}_i \boldsymbol{\gamma}.$$

Here $\mathbf{z}_i \boldsymbol{\gamma}$ denotes the fixed-effect part of the scale model. Thus, mean-only models ask whether predictors explain average EPP, whereas mean-scale models additionally ask whether predictors explain variation in the precision of EPP estimates.

## Prior distributions

The preregistration did not specify prior distributions, so their selection was an analytical decision made for model fitting and is reported explicitly here. The mean-only models used the following `priors_mu` block:

```{r}
#| label: mean-only-priors
#| eval: false

priors_mu <- c(
  prior(normal(0, 2), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "phi")
)
```

Equivalently, the intercept had a $\mathcal{N}(0, 2^2)$ prior, each fixed-effect coefficient had a $\mathcal{N}(0, 1^2)$ prior, each random-effect standard deviation had an $\operatorname{Exponential}(1)$ prior, and the constant beta-binomial precision parameter $\phi$ had an $\operatorname{Exponential}(1)$ prior. The normal priors are weakly informative on the logit scale: they allow a broad range of baseline EPP probabilities and predictor effects while regularising implausibly extreme values; the exponential priors provide weak regularisation for positive scale parameters. For mean-scale models, the priors for the mean and random-effect components were unchanged, while the intercept and coefficients in the $\log(\phi_i)$ component each received $\mathcal{N}(0, 1^2)$ priors.

## Sampler settings

Models were fitted using four independent Markov chains run on four cores. Each chain used 10,000 iterations, of which the first 5,000 were warmup, leaving 20,000 post-warmup draws across chains. The Stan controls were set to `adapt_delta = 0.99` and `max_treedepth = 15`. Model-specific random-number seeds were fixed in the fitting scripts.

## Posterior summaries and inferential interpretation

For each model coefficient $\theta$, we report the posterior mean, the equal-tailed 95% credible interval defined by the 0.025 and 0.975 posterior quantiles, and the posterior probability of its direction. The output columns `prob_positive` and `prob_negative` are, respectively,

$$
\Pr(\theta > 0 \mid y)
\quad\text{and}\quad
\Pr(\theta < 0 \mid y).
$$

The probability of direction is

$$
pd = \max\{\Pr(\theta > 0 \mid y),\Pr(\theta < 0 \mid y)\}.
$$

Thus, `prob_positive` equals $pd$ only when a positive coefficient is the more probable direction; for a predominantly negative coefficient, $pd$ equals `prob_negative`. We use these quantities as continuous summaries of effect direction and uncertainty. We do not apply a posterior-probability, credible-interval, or conventional significance threshold to classify effects as statistically significant or non-significant. Interpretation is based jointly on effect magnitude, the 95% credible interval, and the probability of direction.

MCMC performance is assessed from split $\widehat{R}$, bulk and tail effective sample sizes, divergent transitions, and transitions reaching the maximum tree depth. These diagnostics assess the reliability of posterior computation and are not treated as evidence for or against a biological effect.

## Baseline overall model

The baseline overall model uses only the common grouping structure, without environmental predictors. Its purpose is to provide the preregistered reference point for variance decomposition: the posterior variance attributable to population identity, non-phylogenetic species identity, and phylogenetic species covariance is estimated before adding environmental or methodological moderators. The same random-effect components can then be compared with those from models containing moderators to assess whether the moderators account for variation that was assigned to these grouping levels in the baseline model. The baseline model is therefore not used to test a separate intercept-only biological hypothesis.

Its expected-EPP component is:

$$\operatorname{logit}(\mu_i) =
\alpha +
u_{\operatorname{population}[i]} +
v_{\operatorname{species}[i]} +
w_{\operatorname{phylogeny}[i]}.$$

The precision component is:

$$\log(\phi_i) = \gamma_0.$$

In `brms` notation:

```{r}
#| label: overall-model-formula
#| eval: false

m_overall_population <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    1 +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_model,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log")
)
```

## Baseline variance decomposition

The table below reports posterior summaries for each random-intercept standard deviation, its corresponding variance on the latent logit scale, and its proportion of the summed random-intercept variance. Proportions are calculated within every posterior draw before summarisation, so their uncertainty retains the joint posterior dependence among components. The denominator contains the three grouping-level variances only; it does not include observation-level beta-binomial variation, which does not have a single constant residual variance on the latent scale.

```{r}
#| label: overall-model-variance-components
#| echo: true

if (!is.na(overall_variance_file)) {
  read_csv_safe(overall_variance_file) %>%
    select(
      component,
      variance_posterior_mean,
      variance_lower_95,
      variance_upper_95,
      proportion_posterior_mean,
      proportion_lower_95,
      proportion_upper_95
    ) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
    knitr::kable(
      col.names = c(
        "Component",
        "Variance mean", "Variance 2.5%", "Variance 97.5%",
        "Proportion mean", "Proportion 2.5%", "Proportion 97.5%"
      )
    )
} else {
  cat("Baseline variance-component summary is not available.")
}
```

## Overall model summary

```{r}
#| label: overall-model-result
#| echo: true
#| results: asis

if (!is.na(overall_summary_file)) {
  cat("<details><summary>Show overall model summary</summary>\n\n```\n")
  cat(readLines(overall_summary_file), sep = "\n")
  cat("\n```\n</details>\n")
} else {
  invisible(NULL)
}
```

## Overall model diagnostics

```{r}
#| label: overall-model-diagnostics
#| echo: true

if (!is.na(overall_diagnostics_file)) {
  read_csv_safe(overall_diagnostics_file) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
    knitr::kable()
} else {
  invisible(NULL)
}
```

