---
title: "Main environmental models"
---

## Overview

This chapter defines the final primary environmental model set. The primary model set was organised around three environmental hypotheses.

- The PCA models tested whether broad environmental-geographic gradients were associated with EPP.

- The short-anomaly models tested whether recent deviations from local climatic conditions, measured over the five-year period preceding the breeding record, were associated with EPP.

- The long-anomaly models tested whether deviations from the 1961-1989 climatic baseline were associated with EPP.

Each hypothesis group is presented as a mean-only model, allowing environmental predictors and methodological controls to affect the expected EPP rate while using a single global precision parameter (`phi ~ 1`). Methodological controls are nest box and marker type. All models use the same random-effect structure for population, non-phylogenetic species identity, and phylogenetic species covariance. Exploratory location-scale versions of these models are presented in a separate chapter.

```{r}
#| label: main-env-setup
#| echo: true

library(dplyr)
library(readr)
library(tibble)
library(knitr)

model_dir <- "models/main_environmental_models"

first_existing <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) {
    return(NA_character_)
  }
  existing[1]
}

data_checks_file <- first_existing(c(
  file.path(model_dir, "main_environmental_model_data_checks.csv")
))

fixed_effects_file <- first_existing(c(
  file.path(model_dir, "fixed_effects_main_environmental_models.csv")
))

summary_files <- list(
  pca_mu = first_existing(c(file.path(model_dir, "summary_m_main_pca_mu.txt"))),
  short_mu = first_existing(c(file.path(model_dir, "summary_m_main_short_mu.txt"))),
  long_mu = first_existing(c(file.path(model_dir, "summary_m_main_long_mu.txt")))
)

figure_files <- list(
  pca_effects = "figures/main_model_pca_effects.png",
  pca_mu_forest = "figures/main_pca_mean_forest_plot.png",
  pca_controls = "figures/main_pca_controls_effects.png",
  short_mu_forest = "figures/main_short_mean_forest_plot.png",
  long_mu_forest = "figures/main_long_mean_forest_plot.png"
)

show_saved_summary <- function(summary_file, label) {
  if (!is.na(summary_file) && file.exists(summary_file)) {
    cat("<details><summary>", label, "</summary>\n\n```\n", sep = "")
    cat(readLines(summary_file), sep = "\n")
    cat("\n```\n</details>\n")
  }
}

show_saved_figure <- function(figure_file, missing_message) {
  if (file.exists(figure_file)) {
    knitr::include_graphics(figure_file)
  } else {
    cat(missing_message)
  }
}
```

## Model set

| Hypothesis group | Mean-only model |
|----|----|
| Environmental PCA | `PC1_thermal_geographic + PC2_wet_productive + nest_box + marker_type`; `phi ~ 1` |
| Short anomaly | `tmp_short_anom_5yr_C_first_month_z + pre_short_anom_5yr_mm_first_month_z + abs_latitude_z + distance_to_coast_z + nest_box + marker_type`; `phi ~ 1` |
| Long anomaly | `tmp_long_anom_1961_1989_C_first_month_z + pre_long_anom_1961_1989_mm_first_month_z + abs_latitude_z + distance_to_coast_z + nest_box + marker_type`; `phi ~ 1` |

## Environmental PCA model

```{r}
#| label: main-pca-mu-code
#| eval: false

m_main_pca_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    PC1_thermal_geographic +
    PC2_wet_productive +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1401,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-pca-mu-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$pca_mu, "Show summary for the Environmental PCA mean-only model")
```

```{r}
#| label: main-pca-effects-panel
#| echo: true
#| fig-width: 9
#| fig-height: 4.8

show_saved_figure(
  figure_files$pca_effects,
  "PCA effect plot is not available yet. Render the Figures chapter to create it."
)
```

Lines show posterior mean predicted EPP, shaded ribbons show 95% credible intervals, and pale points show observed population-level EPP values as background context.

```{r}
#| label: main-pca-controls-effects
#| echo: true
#| fig-width: 8
#| fig-height: 3.6

show_saved_figure(
  figure_files$pca_controls,
  ""
)
```

```{r}
#| label: main-pca-mu-forest
#| echo: true
#| fig-width: 8
#| fig-height: 3.2

show_saved_figure(
  figure_files$pca_mu_forest,
  "Forest plot for the Environmental PCA mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Fixed-effect summary for the Environmental PCA mean-only model. Points show posterior estimates and horizontal intervals show 95% credible intervals on the logit scale. Categorical effects are contrasts against the reference groups: nest box = no and marker type = DNA fingerprinting; Nest box compares records with nest-box use to records without nest-box use.


## Short-anomaly model

```{r}
#| label: main-short-mu-code
#| eval: false

m_main_short_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    tmp_short_anom_5yr_C_first_month_z +
    pre_short_anom_5yr_mm_first_month_z +
    abs_latitude_z +
    distance_to_coast_z +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1403,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-short-mu-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$short_mu, "Show summary for the short-anomaly mean-only model")
```

```{r}
#| label: main-short-mu-forest
#| echo: true
#| fig-width: 8
#| fig-height: 3.8

show_saved_figure(
  figure_files$short_mu_forest,
  "Forest plot for the short-anomaly mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Fixed-effect summary for the short-anomaly mean-only model. Points show posterior estimates and horizontal intervals show 95% credible intervals on the logit scale. Categorical effects are contrasts against the reference groups: nest box = no and marker type = DNA fingerprinting; Nest box compares records with nest-box use to records without nest-box use.


## Long-anomaly model

```{r}
#| label: main-long-mu-code
#| eval: false

m_main_long_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    tmp_long_anom_1961_1989_C_first_month_z +
    pre_long_anom_1961_1989_mm_first_month_z +
    abs_latitude_z +
    distance_to_coast_z +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1405,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-long-mu-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$long_mu, "Show summary for the long-anomaly mean-only model")
```

```{r}
#| label: main-long-mu-forest
#| echo: true
#| fig-width: 8
#| fig-height: 3.8

show_saved_figure(
  figure_files$long_mu_forest,
  "Forest plot for the long-anomaly mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Fixed-effect summary for the long-anomaly mean-only model. Points show posterior estimates and horizontal intervals show 95% credible intervals on the logit scale. Categorical effects are contrasts against the reference groups: nest box = no and marker type = DNA fingerprinting; Nest box compares records with nest-box use to records without nest-box use.


## Fixed effects

```{r}
#| label: main-env-fixed-effects
#| echo: true
#| results: asis

if (!is.na(fixed_effects_file)) {
  fixed_effects_table <- read_csv(fixed_effects_file, show_col_types = FALSE) %>%
    mutate(
      model = gsub("Short anomaly", "Short-anomaly", model, fixed = TRUE),
      model = gsub("Long anomaly", "Long-anomaly", model, fixed = TRUE)
    ) %>%
    filter(component == "mu", model %in% c(
      "Environmental PCA mean-only",
      "Short-anomaly mean-only",
      "Long-anomaly mean-only"
    )) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))

  cat("<details><summary>Show fixed-effect table</summary>\n\n")
  cat(knitr::kable(fixed_effects_table, format = "html"))
  cat("\n</details>\n")
}
```





