---
title: "Main environmental models"
---

## Overview

This chapter defines the final primary environmental model set. The primary model set was organised around three environmental hypotheses.

- The PCA models tested whether broad environmental-geographic gradients were associated with EPP.

- The short-anomaly models tested whether recent deviations from local climatic conditions were associated with EPP.

- The long-anomaly models tested whether deviations from the 1961-1989 climatic baseline were associated with EPP.

Each hypothesis group was fitted as a mean-only model and as a location-scale model, allowing environmental predictors to affect the expected EPP rate and, in the mean-scale version, allowing environmental predictors and methodological controls to affect residual dispersion. Methodological controls are `nest_box` and `marker_type`. All models use the same random-effect structure for population, non-phylogenetic species identity, and phylogenetic species covariance.

```{r}
#| label: main-env-setup
#| include: false

library(dplyr)
library(readr)
library(tibble)
library(knitr)

model_dir <- "models/main_environmental_models"

first_existing <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) {
    return(NA_character_)
  }
  existing[1]
}

data_checks_file <- first_existing(c(
  file.path(model_dir, "main_environmental_model_data_checks.csv")
))

fixed_effects_file <- first_existing(c(
  file.path(model_dir, "fixed_effects_main_environmental_models.csv")
))

summary_files <- list(
  pca_mu = first_existing(c(file.path(model_dir, "summary_m_main_pca_mu.txt"))),
  pca_mu_phi = first_existing(c(file.path(model_dir, "summary_m_main_pca_mu_phi.txt"))),
  short_mu = first_existing(c(file.path(model_dir, "summary_m_main_short_mu.txt"))),
  short_mu_phi = first_existing(c(file.path(model_dir, "summary_m_main_short_mu_phi.txt"))),
  long_mu = first_existing(c(file.path(model_dir, "summary_m_main_long_mu.txt"))),
  long_mu_phi = first_existing(c(file.path(model_dir, "summary_m_main_long_mu_phi.txt")))
)

figure_files <- list(
  pca_effects = "figures/main_model_pca_effects.png",
  pca_mu_forest = "figures/main_pca_mean_forest_plot.png",
  short_mu_forest = "figures/main_short_mean_forest_plot.png",
  long_mu_forest = "figures/main_long_mean_forest_plot.png"
)

show_saved_summary <- function(summary_file, label) {
  if (!is.na(summary_file) && file.exists(summary_file)) {
    cat("<details><summary>", label, "</summary>\n\n```\n", sep = "")
    cat(readLines(summary_file), sep = "\n")
    cat("\n```\n</details>\n")
  }
}

show_saved_figure <- function(figure_file, missing_message) {
  if (file.exists(figure_file)) {
    knitr::include_graphics(figure_file)
  } else {
    cat(missing_message)
  }
}
```

## Model set

| Hypothesis group | Mean-only model | Mean-scale model |
|----|----|----|
| Environmental PCA | Mean: `PC1_thermal_geographic + PC2_wet_productive + nest_box + marker_type`; `phi ~ 1` | same mean model; `phi ~ PC1_thermal_geographic + PC2_wet_productive + nest_box + marker_type` |
| Short anomaly | Mean: `tmp_short_anom_5yr_C_first_month_z + pre_short_anom_5yr_mm_first_month_z + abs_latitude_z + distance_to_coast_z + nest_box + marker_type`; `phi ~ 1` | same mean model; `phi ~ tmp_short_anom_5yr_C_first_month_z + pre_short_anom_5yr_mm_first_month_z + nest_box + marker_type` |
| Long anomaly | Mean: `tmp_long_anom_1961_1989_C_first_month_z + pre_long_anom_1961_1989_mm_first_month_z + abs_latitude_z + distance_to_coast_z + nest_box + marker_type`; `phi ~ 1` | same mean model; `phi ~ tmp_long_anom_1961_1989_C_first_month_z + pre_long_anom_1961_1989_mm_first_month_z + nest_box + marker_type` |

## Environmental PCA model

```{r}
#| label: main-pca-mu-code
#| eval: false

m_main_pca_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    PC1_thermal_geographic +
    PC2_wet_productive +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1401,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-pca-mu-summary
#| echo: false
#| results: asis

show_saved_summary(summary_files$pca_mu, "Show summary for the Environmental PCA mean-only model")
```

```{r}
#| label: main-pca-effects-panel
#| echo: false
#| fig-width: 9
#| fig-height: 4.8

show_saved_figure(
  figure_files$pca_effects,
  "PCA effect plot is not available yet. Render the Figures chapter to create it."
)
```

Lines show posterior mean predicted EPP, shaded ribbons show 95% credible intervals, and pale points show observed population-level EPP values as background context.
```{r}
#| label: main-pca-mu-forest
#| echo: false
#| fig-width: 8
#| fig-height: 3.2

show_saved_figure(
  figure_files$pca_mu_forest,
  "Forest plot for the Environmental PCA mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Points show posterior fixed-effect estimates and horizontal intervals show 95% credible intervals on the logit scale.

## Environmental PCA location-scale model

```{r}
#| label: main-pca-mu-phi-code
#| eval: false

m_main_pca_mu_phi <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      PC1_thermal_geographic +
      PC2_wet_productive +
      nest_box +
      marker_type +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat_fit)),

    phi ~
      PC1_thermal_geographic +
      PC2_wet_productive +
      nest_box +
      marker_type
  ),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1402,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-pca-mu-phi-summary
#| echo: false
#| results: asis

show_saved_summary(summary_files$pca_mu_phi, "Show summary for the Environmental PCA location-scale model")
```

## Short-anomaly model

```{r}
#| label: main-short-mu-code
#| eval: false

m_main_short_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    tmp_short_anom_5yr_C_first_month_z +
    pre_short_anom_5yr_mm_first_month_z +
    abs_latitude_z +
    distance_to_coast_z +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1403,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-short-mu-summary
#| echo: false
#| results: asis

show_saved_summary(summary_files$short_mu, "Show summary for the short-anomaly mean-only model")
```

```{r}
#| label: main-short-mu-forest
#| echo: false
#| fig-width: 8
#| fig-height: 3.8

show_saved_figure(
  figure_files$short_mu_forest,
  "Forest plot for the short-anomaly mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Points show posterior fixed-effect estimates and horizontal intervals show 95% credible intervals on the logit scale.

## Short-anomaly location-scale model

```{r}
#| label: main-short-mu-phi-code
#| eval: false

m_main_short_mu_phi <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      tmp_short_anom_5yr_C_first_month_z +
      pre_short_anom_5yr_mm_first_month_z +
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat_fit)),

    phi ~
      tmp_short_anom_5yr_C_first_month_z +
      pre_short_anom_5yr_mm_first_month_z +
      nest_box +
      marker_type
  ),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1404,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-short-mu-phi-summary
#| echo: false
#| results: asis

show_saved_summary(summary_files$short_mu_phi, "Show summary for the short-anomaly location-scale model")
```

## Long-anomaly model

```{r}
#| label: main-long-mu-code
#| eval: false

m_main_long_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    tmp_long_anom_1961_1989_C_first_month_z +
    pre_long_anom_1961_1989_mm_first_month_z +
    abs_latitude_z +
    distance_to_coast_z +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1405,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-long-mu-summary
#| echo: false
#| results: asis

show_saved_summary(summary_files$long_mu, "Show summary for the long-anomaly mean-only model")
```

```{r}
#| label: main-long-mu-forest
#| echo: false
#| fig-width: 8
#| fig-height: 3.8

show_saved_figure(
  figure_files$long_mu_forest,
  "Forest plot for the long-anomaly mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Points show posterior fixed-effect estimates and horizontal intervals show 95% credible intervals on the logit scale.

## Long-anomaly location-scale model

```{r}
#| label: main-long-mu-phi-code
#| eval: false

m_main_long_mu_phi <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      tmp_long_anom_1961_1989_C_first_month_z +
      pre_long_anom_1961_1989_mm_first_month_z +
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat_fit)),

    phi ~
      tmp_long_anom_1961_1989_C_first_month_z +
      pre_long_anom_1961_1989_mm_first_month_z +
      nest_box +
      marker_type
  ),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1406,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-long-mu-phi-summary
#| echo: false
#| results: asis

show_saved_summary(summary_files$long_mu_phi, "Show summary for the long-anomaly location-scale model")
```


## Fixed effects

```{r}
#| label: main-env-fixed-effects
#| echo: false

if (!is.na(fixed_effects_file)) {
  read_csv(fixed_effects_file, show_col_types = FALSE) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
    kable()
} else {
  cat("Fixed-effect summaries for the main environmental models are not available yet.")
}
```

## Predictive support for location-scale models

Pairwise LOO comparisons are used to evaluate whether adding a scale component improves predictive performance relative to the corresponding mean-only model. The table below reports whether the required LOO objects are available for each hypothesis group. Complete pairwise comparisons are shown only when both the mean-only and location-scale LOO objects are present.

```{r}
#| label: main-location-scale-loo-availability
#| echo: false

loo_pairs <- tibble::tribble(
  ~hypothesis_group, ~mean_only_loo, ~location_scale_loo,
  "Environmental PCA", file.path(model_dir, "loo_m_main_pca_mu.rds"), file.path(model_dir, "loo_m_main_pca_mu_phi.rds"),
  "Short anomaly", file.path(model_dir, "loo_m_main_short_mu.rds"), file.path(model_dir, "loo_m_main_short_mu_phi.rds"),
  "Long anomaly", file.path(model_dir, "loo_m_main_long_mu.rds"), file.path(model_dir, "loo_m_main_long_mu_phi.rds")
) %>%
  mutate(
    mean_only_available = file.exists(mean_only_loo),
    location_scale_available = file.exists(location_scale_loo),
    comparison_available = mean_only_available & location_scale_available
  ) %>%
  select(
    `Hypothesis group` = hypothesis_group,
    `Mean-only LOO available` = mean_only_available,
    `Location-scale LOO available` = location_scale_available,
    `Pairwise comparison available` = comparison_available
  )

loo_pairs %>%
  kable()
```

At present, the main environmental model summaries and fitted model objects are available, but the full set of LOO objects required for all three mean-only versus location-scale comparisons is not complete. Once these LOO objects are saved, this section can be expanded into a direct support table for the scale component.





