---
title: "Scale variation across species, populations and years"
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(brms)
library(dplyr)
library(knitr)
```

# Aim

This analysis asks whether residual EPP dispersion differs among species, populations and study years.


::: {.callout-important title="Interpreting Phi"}
In a beta-binomial model, `phi` is a precision parameter. Higher `phi` indicates lower residual dispersion around the mean EPP probability. Lower `phi` indicates greater residual dispersion.
:::

# Load Prepared Data

```{r}
#| label: load-data

dat_model_phylo <- readRDS("data/processed/dat_model_phylo.rds")
```

# Prepare Model Data

```{r}
#| label: prepare-data

complete_model_data <- function(data, vars) {
  data <- data[
    !is.na(data$n_epp_broods) &
      !is.na(data$n_broods_sampled) &
      data$n_broods_sampled > 0 &
      data$n_epp_broods <= data$n_broods_sampled,
  ]

  data <- data[complete.cases(data[, vars]), ]
  droplevels(data)
}

dat_scale <- dat_model_phylo

dat_scale$species_nonphylo <- as.factor(dat_scale$species_nonphylo)
dat_scale$population_id <- as.factor(dat_scale$population_id)

dat_scale$year_of_study_start <- as.numeric(dat_scale$year_of_study_start)
dat_scale$year_of_study_end <- as.numeric(dat_scale$year_of_study_end)
dat_scale$study_year <- ifelse(
  !is.na(dat_scale$year_of_study_end),
  rowMeans(
    cbind(dat_scale$year_of_study_start, dat_scale$year_of_study_end),
    na.rm = TRUE
  ),
  dat_scale$year_of_study_start
)
dat_scale$study_year_factor <- as.factor(round(dat_scale$study_year))

scale_model_vars <- c(
  "n_epp_broods",
  "n_broods_sampled",
  "species_nonphylo",
  "population_id",
  "study_year_factor"
)

dat_scale_model <- complete_model_data(
  dat_scale,
  scale_model_vars
)
```

# Data Support

```{r}
#| label: scale-data-support

scale_data_support <- data.frame(
  quantity = c(
    "Records",
    "Species",
    "Populations",
    "Study years"
  ),
  value = c(
    nrow(dat_scale_model),
    n_distinct(dat_scale_model$species_nonphylo),
    n_distinct(dat_scale_model$population_id),
    n_distinct(dat_scale_model$study_year_factor)
  )
)

kable(scale_data_support)
```

# Priors

```{r}
#| label: priors

priors_scale <- c(
  prior(normal(0, 2), class = "Intercept"),
  prior(exponential(1), class = "sd"),
  prior(normal(0, 2), class = "Intercept", dpar = "phi"),
  prior(exponential(1), class = "sd", dpar = "phi")
)
```

# Scale Model

```{r}
#| label: fit-scale-model-on-totoro
#| eval: false

m_scale_species_population_year <- brm(
  bf(
    n_epp_broods | trials(n_broods_sampled) ~
      1 +
      (1 | mu_species | species_nonphylo) +
      (1 | mu_population | population_id),
    phi ~
      1 +
      (1 | phi_species | species_nonphylo) +
      (1 | phi_population | population_id) +
      (1 | phi_year | study_year_factor)
  ),
  data = dat_scale_model,
  family = beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_scale,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 301,
  save_pars = save_pars(all = TRUE)
)
```

# Model Summary

```{r}
#| label: load-scale-model-summary
#| message: false
#| warning: false

if (file.exists("models/models_scale/m_scale_species_population_year.rds")) {
  m_scale_species_population_year <- readRDS(
    "models/models_scale/m_scale_species_population_year.rds"
  )

  summary(m_scale_species_population_year)
} else {
  cat(
    "Scale model not found: models/models_scale/m_scale_species_population_year.rds\n"
  )
}
```

# Scale Components

```{r}
#| label: scale-components
#| message: false
#| warning: false

if (file.exists("models/models_scale/scale_parameter_components.csv")) {
  scale_parameter_components <- read.csv(
    "models/models_scale/scale_parameter_components.csv"
  )

  kable(scale_parameter_components, digits = 3)
} else {
  cat(
    "Scale component table not found: models/models_scale/scale_parameter_components.csv\n"
  )
}
```

The posterior estimates suggest that residual EPP dispersion can vary at the population, species and study-year levels. However, uncertainty is broad and all credible intervals extend close to zero. This means that the model does not provide strong evidence that scale variation is clearly dominated by one biological level. Population-level dispersion is estimated slightly higher than species- and year-level dispersion, but the difference should be interpreted cautiously.

# Output Files

The corresponding Totoro script is:

```text
scripts/run_scale_parameter_model.R
```

The expected model outputs are:

```text
models/models_scale/m_scale_species_population_year.rds
models/models_scale/summary_m_scale_species_population_year.txt
models/models_scale/scale_model_data_support.csv
models/models_scale/scale_parameter_components.csv
```
