---
title: "Main environmental mean models"
---

## Overview

This chapter defines the final primary environmental model set. The primary model set was organised around three environmental hypotheses.

The preregistration specified direct environmental and geographic predictors rather than PCA axes. Because those predictors are strongly collinear, the PCA specification was added after registration and is reported as the primary multivariate analysis. The registered direct-predictor specification is reported in the following chapter and is used to assess sensitivity to this dimensional-reduction decision.

- The PCA models tested whether broad environmental-geographic gradients were associated with EPP.

- The short-anomaly models tested whether signed recent deviations from local climatic conditions, measured relative to the five years preceding the breeding record, were associated with EPP.

- The long-anomaly models tested whether signed deviations from the 1961-1989 climatic baseline were associated with EPP.

Each hypothesis group is presented as a mean-only model, allowing environmental predictors and methodological controls to affect the expected EPP rate while using a single global precision parameter (`phi ~ 1`). Methodological controls are nest box and marker type. All models use the same random-effect structure for population, non-phylogenetic species identity, and phylogenetic species covariance. Exploratory location-scale versions of these models are presented in a separate chapter.

```{r}
#| label: main-env-setup
#| echo: true

library(dplyr)
library(readr)
library(tibble)
library(knitr)

read_csv_safe <- function(path) {
  tibble::as_tibble(utils::read.csv(
    path, check.names = FALSE, stringsAsFactors = FALSE
  ))
}

model_dir <- "models/main_environmental_models"
anomaly_data_file <- file.path(model_dir, "dat_main_environmental_models.rds")

first_existing <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) {
    return(NA_character_)
  }
  existing[1]
}

data_checks_file <- first_existing(c(
  file.path(model_dir, "main_environmental_model_data_checks.csv")
))

fixed_effects_file <- first_existing(c(
  file.path(model_dir, "fixed_effects_main_environmental_models.csv")
))

diagnostics_file <- first_existing(c(
  file.path(model_dir, "diagnostics_main_environmental_models.csv")
))

main_variance_file <- first_existing(c(
  file.path(model_dir, "variance_components_main_environmental_models.csv")
))

phi_file <- first_existing(c(
  file.path(model_dir, "phi_main_environmental_models.csv")
))

overall_variance_file <- first_existing(c(
  "models/overall_population/overall_population_variance_components.csv"
))

overall_data_checks_file <- first_existing(c(
  "models/overall_population/overall_population_data_checks.csv"
))

if (is.na(fixed_effects_file)) {
  stop("Fixed-effect summaries are required for the model-specific interpretations.")
}

if (!file.exists(anomaly_data_file)) {
  stop("The saved main-model data frame is required to audit the anomaly definitions.")
}

anomaly_data <- readRDS(anomaly_data_file)

anomaly_spec <- tribble(
  ~predictor, ~observed_col, ~baseline_col, ~anomaly_col,
  "Short temperature", "tmp_cru_C_first_month", "tmp_short_base_5yr_C_first_month", "tmp_short_anom_5yr_C_first_month",
  "Short precipitation", "pre_cru_mm_first_month", "pre_short_base_5yr_mm_first_month", "pre_short_anom_5yr_mm_first_month",
  "Long temperature", "tmp_cru_C_first_month", "tmp_long_base_1961_1989_C_first_month", "tmp_long_anom_1961_1989_C_first_month",
  "Long precipitation", "pre_cru_mm_first_month", "pre_long_base_1961_1989_mm_first_month", "pre_long_anom_1961_1989_mm_first_month"
)

required_anomaly_columns <- unique(unlist(anomaly_spec[c(
  "observed_col", "baseline_col", "anomaly_col"
)]))
missing_anomaly_columns <- setdiff(required_anomaly_columns, names(anomaly_data))

if (length(missing_anomaly_columns) > 0L) {
  stop(
    "Columns required for the anomaly audit are missing: ",
    paste(missing_anomaly_columns, collapse = ", ")
  )
}

anomaly_audit <- bind_rows(lapply(seq_len(nrow(anomaly_spec)), function(i) {
  observed <- anomaly_data[[anomaly_spec$observed_col[[i]]]]
  baseline <- anomaly_data[[anomaly_spec$baseline_col[[i]]]]
  anomaly <- anomaly_data[[anomaly_spec$anomaly_col[[i]]]]
  complete <- complete.cases(observed, baseline, anomaly)

  tibble(
    predictor = anomaly_spec$predictor[[i]],
    n_compared = sum(complete),
    n_negative = sum(anomaly[complete] < 0),
    n_zero = sum(anomaly[complete] == 0),
    n_positive = sum(anomaly[complete] > 0),
    max_abs_reconstruction_error = max(abs(
      anomaly[complete] - (observed[complete] - baseline[complete])
    ))
  )
}))

anomaly_tolerance <- 1e-6
if (any(anomaly_audit$max_abs_reconstruction_error > anomaly_tolerance)) {
  stop("A stored climate anomaly is not equal to study-year climate minus its baseline.")
}

main_fixed_effects_results <- read_csv_safe(fixed_effects_file)

effect_row <- function(model_name, predictor_name) {
  result <- main_fixed_effects_results %>%
    filter(model == model_name, predictor == predictor_name)
  if (nrow(result) != 1L) {
    stop(
      "Expected exactly one fixed-effect row for ", model_name,
      " / ", predictor_name, "."
    )
  }
  result
}

pca_pc1 <- effect_row("Environmental PCA mean-only", "PC1 thermal-geographic")
pca_pc2 <- effect_row("Environmental PCA mean-only", "PC2 wet-productivity")
short_temperature <- effect_row("Short-anomaly mean-only", "Short temperature anomaly")
short_precipitation <- effect_row("Short-anomaly mean-only", "Short precipitation anomaly")
long_temperature <- effect_row("Long-anomaly mean-only", "Long temperature anomaly")
long_precipitation <- effect_row("Long-anomaly mean-only", "Long precipitation anomaly")

clear_effects <- main_fixed_effects_results %>%
  filter(
    model %in% c(
      "Environmental PCA mean-only",
      "Short-anomaly mean-only",
      "Long-anomaly mean-only"
    ),
    lower > 0 | upper < 0
  )

if (
  nrow(clear_effects) != 1L ||
  clear_effects$model[[1]] != "Environmental PCA mean-only" ||
  clear_effects$predictor[[1]] != "PC1 thermal-geographic"
) {
  stop("The written fixed-effect interpretation no longer matches the saved posterior intervals.")
}

summary_files <- list(
  pca_mu = first_existing(c(file.path(model_dir, "summary_m_main_pca_mu.txt"))),
  short_mu = first_existing(c(file.path(model_dir, "summary_m_main_short_mu.txt"))),
  long_mu = first_existing(c(file.path(model_dir, "summary_m_main_long_mu.txt")))
)

figure_files <- list(
  pca_effects = "figures/main_model_pca_effects.png",
  pca_mu_forest = "figures/main_pca_mean_forest_plot.png",
  pca_controls = "figures/main_pca_controls_effects.png",
  short_mu_forest = "figures/main_short_mean_forest_plot.png",
  long_mu_forest = "figures/main_long_mean_forest_plot.png"
)

show_saved_summary <- function(summary_file, label) {
  if (!is.na(summary_file) && file.exists(summary_file)) {
    cat("<details><summary>", label, "</summary>\n\n```\n", sep = "")
    cat(readLines(summary_file), sep = "\n")
    cat("\n```\n</details>\n")
  }
}

show_saved_figure <- function(figure_file, missing_message) {
  if (file.exists(figure_file)) {
    knitr::include_graphics(figure_file)
  } else {
    cat(missing_message)
  }
}
```

## Climate anomaly definitions {#sec-climate-anomaly-definitions}

The anomaly models use the CRU value for **one calendar month only**: the first reported breeding-season month (`climate_month`) in the study year (`climate_year`). They do not average climate over the full breeding window. At locality $\ell$, in study year $y$ and first breeding month $m$, let $X_{\ell,y,m}$ denote either monthly mean temperature (degrees C) or monthly total precipitation (mm). The short-term baseline is the mean for the same locality and calendar month over the five years immediately preceding the study year:

$$
\bar X^{\mathrm{short}}_{\ell,y,m}
= \frac{1}{5}\sum_{k=y-5}^{y-1} X_{\ell,k,m},
\qquad
A^{\mathrm{short}}_{\ell,y,m}
= X_{\ell,y,m}-\bar X^{\mathrm{short}}_{\ell,y,m}.
$$

The long-term baseline is the mean for the same locality and calendar month over 1961--1989:

$$
\bar X^{\mathrm{long}}_{\ell,m}
= \frac{1}{29}\sum_{k=1961}^{1989} X_{\ell,k,m},
\qquad
A^{\mathrm{long}}_{\ell,y,m}
= X_{\ell,y,m}-\bar X^{\mathrm{long}}_{\ell,m}.
$$

Thus, all four modelled anomalies are **signed differences, not absolute deviations**. A positive temperature anomaly means that the first breeding month was warmer than its local baseline and a negative value means that it was cooler. A positive precipitation anomaly means wetter conditions and a negative value means drier conditions. The corresponding full-season variables in the dataset were constructed by the same signed subtraction across the full breeding window, but they are not used in these two primary anomaly models.

After complete-case filtering, each signed anomaly was standardized within the common 441-record model frame as

$$
A_z = \frac{A-\bar A_{\mathcal D}}{s_{\mathcal D}(A)},
$$

where $\mathcal D$ is the main-model data frame. No absolute-value transformation was applied before or after standardization. Consequently, a positive regression coefficient represents increasing EPP as the anomaly becomes more positive (warmer or wetter relative to baseline); it does **not** represent an association with the magnitude of climatic departure irrespective of direction. Models using $|A|$ would test a different hypothesis and were not fitted here, so their interpretation is not interchangeable with that of the earlier absolute-deviation analysis.

The following executable audit reconstructs every stored anomaly in the model frame from its study-year value and baseline. The presence of both negative and positive values, together with negligible reconstruction error for $A=X-\bar X$, verifies the signed definition rather than relying on the column names.

```{r}
#| label: main-anomaly-definition-audit
#| echo: true

anomaly_audit %>%
  mutate(
    max_abs_reconstruction_error = format(
      max_abs_reconstruction_error,
      scientific = TRUE,
      digits = 2
    )
  ) %>%
  rename(
    Predictor = predictor,
    `Records compared` = n_compared,
    Negative = n_negative,
    Zero = n_zero,
    Positive = n_positive,
    `Max |stored - (observed - baseline)|` = max_abs_reconstruction_error
  ) %>%
  knitr::kable(align = c("l", "r", "r", "r", "r", "r"))
```

## Model set

| Hypothesis group | Mean-only model |
|----|----|
| Environmental PCA | `PC1_thermal_geographic + PC2_wet_productive + nest_box + marker_type`; `phi ~ 1` |
| Short anomaly | `tmp_short_anom_5yr_C_first_month_z + pre_short_anom_5yr_mm_first_month_z + abs_latitude_z + distance_to_coast_z + nest_box + marker_type`; `phi ~ 1` |
| Long anomaly | `tmp_long_anom_1961_1989_C_first_month_z + pre_long_anom_1961_1989_mm_first_month_z + abs_latitude_z + distance_to_coast_z + nest_box + marker_type`; `phi ~ 1` |

## Environmental PCA model

```{r}
#| label: main-pca-mu-code
#| eval: false

m_main_pca_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    PC1_thermal_geographic +
    PC2_wet_productive +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1401,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-pca-mu-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$pca_mu, "Show summary for the Environmental PCA mean-only model")
```

```{r}
#| label: main-pca-effects-panel
#| echo: true
#| fig-width: 9
#| fig-height: 4.8

show_saved_figure(
  figure_files$pca_effects,
  "PCA effect plot is not available yet. Render the Figures chapter to create it."
)
```

Lines show posterior mean predicted EPP, shaded ribbons show 95% credible intervals, and pale points show observed population-level EPP values as background context.

```{r}
#| label: main-pca-controls-effects
#| echo: true
#| fig-width: 8
#| fig-height: 3.6

show_saved_figure(
  figure_files$pca_controls,
  ""
)
```

```{r}
#| label: main-pca-mu-forest
#| echo: true
#| fig-width: 8
#| fig-height: 3.2

show_saved_figure(
  figure_files$pca_mu_forest,
  "Forest plot for the Environmental PCA mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Fixed-effect summary for the Environmental PCA mean-only model. Points show posterior estimates and horizontal intervals show 95% credible intervals on the logit scale. Categorical effects are contrasts against the reference groups: nest box = no and marker type = DNA fingerprinting; Nest box compares records with nest-box use to records without nest-box use.

PC1 was the only fixed effect whose 95% credible interval excluded zero, although only narrowly: its posterior mean was `r sprintf("%.3f", pca_pc1$estimate)` on the logit scale (95% CrI [`r sprintf("%.3f", pca_pc1$lower)`, `r sprintf("%.3f", pca_pc1$upper)`]; probability of a positive direction = `r sprintf("%.3f", pca_pc1$prob_positive)`). This supports a small positive association between EPP and the composite thermal-geographic gradient, not any single constituent of that gradient. PC2 was `r sprintf("%.3f", pca_pc2$estimate)` (95% CrI [`r sprintf("%.3f", pca_pc2$lower)`, `r sprintf("%.3f", pca_pc2$upper)`]), and all methodological-control intervals also included zero.


## Short-anomaly model

```{r}
#| label: main-short-mu-code
#| eval: false

m_main_short_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    tmp_short_anom_5yr_C_first_month_z +
    pre_short_anom_5yr_mm_first_month_z +
    abs_latitude_z +
    distance_to_coast_z +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1403,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-short-mu-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$short_mu, "Show summary for the short-anomaly mean-only model")
```

```{r}
#| label: main-short-mu-forest
#| echo: true
#| fig-width: 8
#| fig-height: 3.8

show_saved_figure(
  figure_files$short_mu_forest,
  "Forest plot for the short-anomaly mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Fixed-effect summary for the short-anomaly mean-only model. Points show posterior estimates and horizontal intervals show 95% credible intervals on the logit scale. Categorical effects are contrasts against the reference groups: nest box = no and marker type = DNA fingerprinting; Nest box compares records with nest-box use to records without nest-box use.

The short temperature-anomaly coefficient was `r sprintf("%.3f", short_temperature$estimate)` on the logit scale (95% CrI [`r sprintf("%.3f", short_temperature$lower)`, `r sprintf("%.3f", short_temperature$upper)`]), and the short precipitation-anomaly coefficient was `r sprintf("%.3f", short_precipitation$estimate)` (95% CrI [`r sprintf("%.3f", short_precipitation$lower)`, `r sprintf("%.3f", short_precipitation$upper)`]). These coefficients compare increasingly warm or wet first breeding months with increasingly cool or dry months relative to the preceding five-year baseline; they do not measure the effect of absolute climatic departure. Both estimates were small and both intervals included zero. No other fixed effect in this model had a 95% credible interval excluding zero, so the model provides no clearly resolved directional association with short-term climatic anomalies, geography, nest-box use, or marker type.


## Long-anomaly model

```{r}
#| label: main-long-mu-code
#| eval: false

m_main_long_mu <- brms::brm(
  n_epp_broods | trials(n_broods_sampled) ~
    tmp_long_anom_1961_1989_C_first_month_z +
    pre_long_anom_1961_1989_mm_first_month_z +
    abs_latitude_z +
    distance_to_coast_z +
    nest_box +
    marker_type +
    (1 | population_id) +
    (1 | species_nonphylo) +
    (1 | gr(species_phylo, cov = phylo_mat_fit)),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1405,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-long-mu-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$long_mu, "Show summary for the long-anomaly mean-only model")
```

```{r}
#| label: main-long-mu-forest
#| echo: true
#| fig-width: 8
#| fig-height: 3.8

show_saved_figure(
  figure_files$long_mu_forest,
  "Forest plot for the long-anomaly mean-only model is not available yet. Render the Figures chapter to create it."
)
```

Fixed-effect summary for the long-anomaly mean-only model. Points show posterior estimates and horizontal intervals show 95% credible intervals on the logit scale. Categorical effects are contrasts against the reference groups: nest box = no and marker type = DNA fingerprinting; Nest box compares records with nest-box use to records without nest-box use.

The long temperature-anomaly coefficient was `r sprintf("%.3f", long_temperature$estimate)` on the logit scale (95% CrI [`r sprintf("%.3f", long_temperature$lower)`, `r sprintf("%.3f", long_temperature$upper)`]), and the long precipitation-anomaly coefficient was `r sprintf("%.3f", long_precipitation$estimate)` (95% CrI [`r sprintf("%.3f", long_precipitation$lower)`, `r sprintf("%.3f", long_precipitation$upper)`]). These are directional warm-versus-cool and wet-versus-dry contrasts relative to 1961--1989, not effects of absolute departure. Both intervals included zero, as did the intervals for latitude, distance from the coast, and the methodological controls. Thus, this model provides no clearly resolved directional association between EPP and anomalies relative to the 1961--1989 climatic baseline.


## Fixed effects

```{r}
#| label: main-env-fixed-effects
#| echo: true
#| results: asis

if (!is.na(fixed_effects_file)) {
  fixed_effects_table <- main_fixed_effects_results %>%
    mutate(
      model = gsub("Short anomaly", "Short-anomaly", model, fixed = TRUE),
      model = gsub("Long anomaly", "Long-anomaly", model, fixed = TRUE)
    ) %>%
    filter(component == "mu", model %in% c(
      "Environmental PCA mean-only",
      "Short-anomaly mean-only",
      "Long-anomaly mean-only"
    )) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))

  cat("<details><summary>Show fixed-effect table</summary>\n\n")
  cat(knitr::kable(fixed_effects_table, format = "html"))
  cat("\n</details>\n")
}
```

## Beta-binomial precision ($\phi$)

The beta-binomial precision parameter quantifies residual extra-binomial heterogeneity after conditioning on the fixed effects and the three random-intercept components. Larger $\phi$ means less extra-binomial variation, with the ordinary binomial approached as $\phi$ increases. The corresponding within-record correlation among Bernoulli trials is $\rho = 1/(\phi + 1)$.

```{r}
#| label: main-env-phi
#| echo: true

if (is.na(phi_file)) {
  stop("The beta-binomial precision summaries are missing.")
}

phi_results <- read_csv_safe(phi_file) %>%
  mutate(
    model_label = factor(model_label, levels = c(
      "Environmental PCA", "Short anomaly", "Long anomaly"
    ))
  ) %>%
  arrange(model_label)

if (nrow(phi_results) != 3L) {
  stop("The phi summary must contain exactly the three primary models.")
}

phi_table <- phi_results %>%
  transmute(
    Model = model_label,
    `phi posterior mean [95% CrI]` = sprintf(
      "%.2f [%.2f, %.2f]",
      phi_posterior_mean, phi_lower_95, phi_upper_95
    ),
    `Residual rho [95% CrI]` = sprintf(
      "%.3f [%.3f, %.3f]",
      residual_rho_posterior_mean,
      residual_rho_lower_95,
      residual_rho_upper_95
    )
  )

kable(phi_table)
```

The three models produced nearly identical $\phi$ estimates (`r sprintf("%.2f", min(phi_results$phi_posterior_mean))`--`r sprintf("%.2f", max(phi_results$phi_posterior_mean))`), corresponding to residual $\rho$ values of approximately `r sprintf("%.3f", mean(phi_results$residual_rho_posterior_mean))`. Thus, appreciable extra-binomial variation remained after accounting for the environmental predictors, methodological controls, population identity, species identity, and phylogeny; none of the three environmental specifications materially changed this residual dispersion.

## MCMC diagnostics

The table reports the maximum rank-normalised $\hat{R}$ and the minimum bulk and tail effective sample sizes across all monitored posterior parameters in each of the three primary mean-only models. Divergences and transitions reaching the configured maximum tree depth are reported across all four chains.

```{r}
#| label: main-env-mcmc-diagnostics
#| echo: true

if (is.na(diagnostics_file)) {
  stop("Main-model diagnostics are missing.")
}

diagnostics_table <- read_csv_safe(diagnostics_file) %>%
  filter(model %in% c("m_main_pca_mu", "m_main_short_mu", "m_main_long_mu")) %>%
  mutate(
    Model = recode(
      model,
      m_main_pca_mu = "Environmental PCA",
      m_main_short_mu = "Short anomaly",
      m_main_long_mu = "Long anomaly"
    ),
    Model = factor(Model, levels = c(
      "Environmental PCA", "Short anomaly", "Long anomaly"
    ))
  ) %>%
  arrange(Model) %>%
  transmute(
    Model,
    `Maximum R-hat` = sprintf("%.3f", max_rhat),
    `Minimum bulk ESS` = round(min_bulk_ess),
    `Minimum tail ESS` = round(min_tail_ess),
    Divergences = divergent_transitions,
    `Maximum-treedepth hits` = treedepth_hits
  )

if (nrow(diagnostics_table) != 3L) {
  stop("Diagnostics must contain exactly the three primary mean-only models.")
}

kable(diagnostics_table)
```

All three models had $\hat{R}$ values close to 1, effective sample sizes exceeding 900, no divergent transitions, and no maximum-treedepth hits.

## Random-effect variance components

The preregistered variance-component comparison is shown below for the baseline model without moderators and for each of the three primary environmental models. Posterior standard deviations are on the logit scale. The reported proportion is the share of the summed population, non-phylogenetic species, and phylogenetic species random-intercept variance; it is not a residual or observation-scale $R^2$.

```{r}
#| label: main-env-variance-components
#| echo: true

required_variance_files <- c(main_variance_file, overall_variance_file)
if (any(is.na(required_variance_files))) {
  stop("Variance-component summaries are missing for the baseline or main models.")
}
if (is.na(data_checks_file) || is.na(overall_data_checks_file)) {
  stop("Data-check summaries are required to report model-frame sizes.")
}

main_n_records <- read_csv_safe(data_checks_file)$n_records[[1]]
overall_n_records <- read_csv_safe(overall_data_checks_file)$n_records[[1]]

main_variance <- read_csv_safe(main_variance_file) %>%
  mutate(n_records = main_n_records)

overall_variance <- read_csv_safe(overall_variance_file) %>%
  mutate(
    model_label = "Baseline (no moderators)",
    n_records = overall_n_records
  )

variance_component_table <- bind_rows(overall_variance, main_variance) %>%
  mutate(
    model_label = factor(model_label, levels = c(
      "Baseline (no moderators)",
      "Environmental PCA",
      "Short anomaly",
      "Long anomaly"
    )),
    component = factor(component, levels = c(
      "Species (phylogenetic)",
      "Species (non-phylogenetic)",
      "Population"
    ))
  ) %>%
  arrange(model_label, component) %>%
  transmute(
    Model = model_label,
    Records = n_records,
    Component = component,
    `Posterior SD [95% CrI]` = sprintf(
      "%.3f [%.3f, %.3f]",
      sd_posterior_mean, sd_lower_95, sd_upper_95
    ),
    `Share of summed random-intercept variance [95% CrI]` = sprintf(
      "%.3f [%.3f, %.3f]",
      proportion_posterior_mean,
      proportion_lower_95,
      proportion_upper_95
    )
  )

kable(variance_component_table)
```

Across all three environmental models, the phylogenetic component had the largest posterior mean SD (`r sprintf("%.3f", min(main_variance$sd_posterior_mean[main_variance$component == "Species (phylogenetic)"]))`--`r sprintf("%.3f", max(main_variance$sd_posterior_mean[main_variance$component == "Species (phylogenetic)"]))`), followed by the non-phylogenetic species component (`r sprintf("%.3f", min(main_variance$sd_posterior_mean[main_variance$component == "Species (non-phylogenetic)"]))`--`r sprintf("%.3f", max(main_variance$sd_posterior_mean[main_variance$component == "Species (non-phylogenetic)"]))`) and the population component (`r sprintf("%.3f", min(main_variance$sd_posterior_mean[main_variance$component == "Population"]))`--`r sprintf("%.3f", max(main_variance$sd_posterior_mean[main_variance$component == "Population"]))`). The credible intervals are broad and overlap, so this ordering describes posterior means rather than a sharply resolved difference between components.

The baseline model used `r overall_n_records` records, whereas the environmental models used the common complete-case frame of `r main_n_records` records. Consequently, the side-by-side comparison documents the stability and ordering of the variance components, but differences between the baseline and moderator models cannot be attributed solely to the moderators. A strict variance-reduction comparison would require fitting the no-moderator model to the same `r main_n_records` records.





