---
title: "Environmental predictors of EPP dispersion"
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(brms)
library(knitr)
```

# Aim

This page documents exploratory distributional models that test whether environmental predictors are associated with residual dispersion in extra-pair paternity (EPP).

The three models mirror the main environmental models:

1. **Absolute climate and NDVI**
2. **Short-term climatic anomalies and NDVI**
3. **Long-term climatic anomalies and NDVI**

In each model, the mean EPP probability (`mu`) and the beta-binomial precision parameter (`phi`) are modelled with the same fixed predictors. This keeps the dispersion models directly comparable to the main environmental models.

::: {.callout-caution title="Interpreting Phi"}
In a beta-binomial model, `phi` is a precision parameter. Positive coefficients for `phi` indicate higher precision and therefore lower residual dispersion. Negative coefficients indicate lower precision and therefore greater residual dispersion.
:::

# Load Prepared Data

```{r}
#| label: load-data

dat_model_phylo <- readRDS("data/processed/dat_model_phylo.rds")
phylo_mat <- readRDS("data/processed/phylo_mat.rds")
```

# Prepare Model Data

```{r}
#| label: prepare-data

z <- function(x) {
  as.numeric(scale(x))
}

complete_model_data <- function(data, vars) {
  data <- data[
    !is.na(data$n_epp_broods) &
      !is.na(data$n_broods_sampled) &
      data$n_broods_sampled > 0 &
      data$n_epp_broods <= data$n_broods_sampled,
  ]

  data <- data[complete.cases(data[, vars]), ]
  droplevels(data)
}

dat_env <- dat_model_phylo

dat_env$nest_box <- as.factor(dat_env$nest_box)
dat_env$marker_type <- as.factor(dat_env$marker_type)
dat_env$population_id <- as.factor(dat_env$population_id)
dat_env$species_nonphylo <- as.factor(dat_env$species_nonphylo)
dat_env$species_phylo <- as.factor(dat_env$species_phylo)

dat_env$abs_latitude_z <- z(dat_env$abs_latitude)
dat_env$distance_to_coast_z <- z(dat_env$distance_to_coast_km)

dat_env$tmp_cru_C_first_month_z <- z(dat_env$tmp_cru_C_first_month)
dat_env$pre_cru_mm_first_month_log_z <- z(log1p(dat_env$pre_cru_mm_first_month))
dat_env$dtr_cru_C_first_month_z <- z(dat_env$dtr_cru_C_first_month)
dat_env$ndvi_first_month_median_30km_z <- z(dat_env$ndvi_first_month_median_30km)

dat_env$tmp_short_anom_5yr_C_first_month_z <- z(dat_env$tmp_short_anom_5yr_C_first_month)
dat_env$pre_short_anom_5yr_mm_first_month_z <- z(dat_env$pre_short_anom_5yr_mm_first_month)

dat_env$tmp_long_anom_1961_1989_C_first_month_z <- z(dat_env$tmp_long_anom_1961_1989_C_first_month)
dat_env$pre_long_anom_1961_1989_mm_first_month_z <- z(dat_env$pre_long_anom_1961_1989_mm_first_month)
```

# Complete Datasets

```{r}
#| label: complete-datasets

common_vars <- c(
  "n_epp_broods",
  "n_broods_sampled",
  "abs_latitude_z",
  "distance_to_coast_z",
  "nest_box",
  "marker_type",
  "population_id",
  "species_nonphylo",
  "species_phylo"
)

absolute_vars <- c(
  common_vars,
  "tmp_cru_C_first_month_z",
  "pre_cru_mm_first_month_log_z",
  "dtr_cru_C_first_month_z",
  "ndvi_first_month_median_30km_z"
)

short_vars <- c(
  common_vars,
  "tmp_short_anom_5yr_C_first_month_z",
  "pre_short_anom_5yr_mm_first_month_z",
  "dtr_cru_C_first_month_z",
  "ndvi_first_month_median_30km_z"
)

long_vars <- c(
  common_vars,
  "tmp_long_anom_1961_1989_C_first_month_z",
  "pre_long_anom_1961_1989_mm_first_month_z",
  "dtr_cru_C_first_month_z",
  "ndvi_first_month_median_30km_z"
)

dat_env_absolute <- complete_model_data(dat_env, absolute_vars)
dat_env_short <- complete_model_data(dat_env, short_vars)
dat_env_long <- complete_model_data(dat_env, long_vars)

phi_model_data_checks <- data.frame(
  model = c(
    "absolute climate phi",
    "short anomalies phi",
    "long anomalies phi"
  ),
  records = c(
    nrow(dat_env_absolute),
    nrow(dat_env_short),
    nrow(dat_env_long)
  ),
  species = c(
    length(unique(dat_env_absolute$species_phylo)),
    length(unique(dat_env_short$species_phylo)),
    length(unique(dat_env_long$species_phylo))
  ),
  populations = c(
    length(unique(dat_env_absolute$population_id)),
    length(unique(dat_env_short$population_id)),
    length(unique(dat_env_long$population_id))
  )
)

kable(phi_model_data_checks)
```

# Priors

```{r}
#| label: priors

priors_env_phi <- c(
  prior(normal(0, 2), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(normal(0, 2), class = "Intercept", dpar = "phi"),
  prior(normal(0, 1), class = "b", dpar = "phi")
)
```

# Model 1: Absolute Climate Phi Model

```{r}
#| label: fit-absolute-phi-on-totoro
#| eval: false

m_env_absolute_phi <- brm(
  bf(
    n_epp_broods | trials(n_broods_sampled) ~
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      tmp_cru_C_first_month_z +
      pre_cru_mm_first_month_log_z +
      dtr_cru_C_first_month_z +
      ndvi_first_month_median_30km_z +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat)),
    phi ~
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      tmp_cru_C_first_month_z +
      pre_cru_mm_first_month_log_z +
      dtr_cru_C_first_month_z +
      ndvi_first_month_median_30km_z
  ),
  data = dat_env_absolute,
  data2 = list(phylo_mat = phylo_mat),
  family = beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_env_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 241,
  save_pars = save_pars(all = TRUE)
)
```

## Summary

```{r}
#| label: load-summary-absolute-phi
#| message: false
#| warning: false

if (file.exists("models/m_env_absolute_phi.rds")) {
  m_env_absolute_phi <- readRDS("models/m_env_absolute_phi.rds")
  summary(m_env_absolute_phi)
} else {
  cat("Model file not found: models/m_env_absolute_phi.rds\n")
}
```

# Model 2: Short-Term Anomaly Phi Model

```{r}
#| label: fit-short-phi-on-totoro
#| eval: false

m_env_short_phi <- brm(
  bf(
    n_epp_broods | trials(n_broods_sampled) ~
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      tmp_short_anom_5yr_C_first_month_z +
      pre_short_anom_5yr_mm_first_month_z +
      dtr_cru_C_first_month_z +
      ndvi_first_month_median_30km_z +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat)),
    phi ~
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      tmp_short_anom_5yr_C_first_month_z +
      pre_short_anom_5yr_mm_first_month_z +
      dtr_cru_C_first_month_z +
      ndvi_first_month_median_30km_z
  ),
  data = dat_env_short,
  data2 = list(phylo_mat = phylo_mat),
  family = beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_env_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 251,
  save_pars = save_pars(all = TRUE)
)
```

## Summary

```{r}
#| label: load-summary-short-phi
#| message: false
#| warning: false

if (file.exists("models/m_env_short_phi.rds")) {
  m_env_short_phi <- readRDS("models/m_env_short_phi.rds")
  summary(m_env_short_phi)
} else {
  cat("Model file not found: models/m_env_short_phi.rds\n")
}
```

# Model 3: Long-Term Anomaly Phi Model

```{r}
#| label: fit-long-phi-on-totoro
#| eval: false

m_env_long_phi <- brm(
  bf(
    n_epp_broods | trials(n_broods_sampled) ~
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      tmp_long_anom_1961_1989_C_first_month_z +
      pre_long_anom_1961_1989_mm_first_month_z +
      dtr_cru_C_first_month_z +
      ndvi_first_month_median_30km_z +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat)),
    phi ~
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      tmp_long_anom_1961_1989_C_first_month_z +
      pre_long_anom_1961_1989_mm_first_month_z +
      dtr_cru_C_first_month_z +
      ndvi_first_month_median_30km_z
  ),
  data = dat_env_long,
  data2 = list(phylo_mat = phylo_mat),
  family = beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_env_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 261,
  save_pars = save_pars(all = TRUE)
)
```

## Summary

```{r}
#| label: load-summary-long-phi
#| message: false
#| warning: false

if (file.exists("models/m_env_long_phi.rds")) {
  m_env_long_phi <- readRDS("models/m_env_long_phi.rds")
  summary(m_env_long_phi)
} else {
  cat("Model file not found: models/m_env_long_phi.rds\n")
}
```

# Model Comparison

```{r}
#| label: loo-phi-models
#| eval: false

loo_env_absolute_phi <- loo(m_env_absolute_phi)
loo_env_short_phi <- loo(m_env_short_phi)
loo_env_long_phi <- loo(m_env_long_phi)

loo_comparison_environmental_phi <- loo_compare(
  loo_env_absolute_phi,
  loo_env_short_phi,
  loo_env_long_phi
)

```

```{r}
#| label: load-loo-comparison-phi
#| message: false
#| warning: false

if (file.exists("models/loo_comparison_environmental_phi.rds")) {
  loo_comparison_environmental_phi <- readRDS(
    "models/loo_comparison_environmental_phi.rds"
  )

  loo_comparison_environmental_phi
} else {
  cat("LOO comparison file not found: models/loo_comparison_environmental_phi.rds\n")
}
```

# Output Files

The expected model outputs are:

```text
models/m_env_absolute_phi.rds
models/m_env_short_phi.rds
models/m_env_long_phi.rds
models/loo_comparison_environmental_phi.rds
```
