---
title: "Confirmatory temperature-anomaly × latitude models"
---

## Overview

This chapter presents a confirmatory re-test of a previously reported interaction: the association between temperature anomaly and extra-pair paternity (EPP) was reported to become less positive with increasing absolute latitude. The earlier result came from a different dataset and non-phylogenetic linear mixed models. Here, the hypothesis is tested in the current beta-binomial phylogenetic framework. Only the two prespecified temperature-anomaly × absolute-latitude interactions are fitted; this is not an exploratory interaction search.

```{r}
#| label: interaction-setup
#| echo: true

library(dplyr)
library(readr)
library(knitr)

read_csv_safe <- function(path) {
  tibble::as_tibble(utils::read.csv(
    path, check.names = FALSE, stringsAsFactors = FALSE
  ))
}

model_dir <- "models/interaction_models"

first_existing <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) return(NA_character_)
  existing[1]
}

show_saved_summary <- function(summary_file, label) {
  if (!is.na(summary_file) && file.exists(summary_file)) {
    cat("<details><summary>", label, "</summary>\n\n```\n", sep = "")
    cat(readLines(summary_file), sep = "\n")
    cat("\n```\n</details>\n")
  } else {
    cat("Saved model summary is not available yet.")
  }
}

show_saved_figure <- function(figure_file, missing_message) {
  if (file.exists(figure_file)) knitr::include_graphics(figure_file) else cat(missing_message)
}

coverage_file <- first_existing(file.path(model_dir, "latitudinal_coverage.csv"))
diagnostics_file <- first_existing(file.path(model_dir, "diagnostics_interaction_models.csv"))
effects_file <- first_existing(file.path(model_dir, "fixed_effects_interaction_models.csv"))
directional_file <- first_existing(file.path(model_dir, "directional_interaction_summary.csv"))
loo_file <- first_existing(file.path(model_dir, "loo_compare_interaction_models.csv"))
pareto_file <- first_existing(file.path(model_dir, "loo_diagnostics_interaction_models.csv"))
prediction_scale_file <- first_existing(file.path(model_dir, "interaction_prediction_scale.csv"))
pareto_audit_dir <- "models/pareto_k_audit"
pareto_audit_summary_file <- first_existing(file.path(pareto_audit_dir, "pareto_k_model_summary.csv"))
pareto_audit_high_file <- first_existing(file.path(pareto_audit_dir, "pareto_k_unique_high_observations.csv"))
pareto_audit_reference_file <- first_existing(file.path(pareto_audit_dir, "pareto_k_overall_reference.csv"))
```

## Latitudinal coverage

The interaction is identifiable only to the extent that the fitted model frame covers low latitudes. The table below is calculated from the exact 441-row `dat_fit` object used for fitting, rather than from the raw database. The number of low-latitude records, populations, and species should therefore be treated as an explicit caveat when interpreting the interaction.

```{r}
#| label: interaction-coverage
#| echo: true
if (!is.na(coverage_file)) {
  kable(read_csv_safe(coverage_file), digits = 2)
} else {
  cat("Latitudinal coverage table will appear after the fitting script has run.")
}
```

## Model set

| Confirmatory model | Prespecified interaction | Other covariates |
|---|---|---|
| Short anomaly × latitude | `tmp_short_anom_5yr_C_first_month_z * abs_latitude_z` | short precipitation anomaly, distance to coast, nest box, marker type |
| Long anomaly × latitude | `tmp_long_anom_1961_1989_C_first_month_z * abs_latitude_z` | long precipitation anomaly, distance to coast, nest box, marker type |

Both models use a beta-binomial response, `phi ~ 1`, and the same population, non-phylogenetic species, and phylogenetic species random intercepts as the main models.

## Short-anomaly interaction

```{r}
#| label: interaction-short-code
#| eval: false
m_interaction_short_latitude <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      tmp_short_anom_5yr_C_first_month_z * abs_latitude_z +
      pre_short_anom_5yr_mm_first_month_z + distance_to_coast_z +
      nest_box + marker_type + (1 | population_id) + (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat_fit)),
    phi ~ 1
  ),
  data = dat_fit, data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu, chains = 4, cores = 4, iter = 10000, warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15), seed = 1407,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: interaction-short-summary
#| results: asis
show_saved_summary(first_existing(file.path(model_dir, "summary_m_interaction_short_latitude.txt")),
                   "Show summary for the short-anomaly × latitude model")
```

## Long-anomaly interaction

```{r}
#| label: interaction-long-code
#| eval: false
m_interaction_long_latitude <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      tmp_long_anom_1961_1989_C_first_month_z * abs_latitude_z +
      pre_long_anom_1961_1989_mm_first_month_z + distance_to_coast_z +
      nest_box + marker_type + (1 | population_id) + (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat_fit)),
    phi ~ 1
  ),
  data = dat_fit, data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu, chains = 4, cores = 4, iter = 10000, warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15), seed = 1409,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: interaction-long-summary
#| results: asis
show_saved_summary(first_existing(file.path(model_dir, "summary_m_interaction_long_latitude.txt")),
                   "Show summary for the long-anomaly × latitude model")
```

## Directional posterior test

The prespecified direction from the earlier analysis is a negative interaction coefficient. Accordingly, the table reports the posterior mean, 95% credible interval, and posterior probability that each interaction is below zero.

```{r}
#| label: interaction-directional
if (!is.na(directional_file)) kable(read_csv_safe(directional_file), digits = 3)
```

Neither confirmatory interaction was supported. The short-term interaction was
$-0.020$ (95% CrI $-0.115$ to $0.075$; posterior probability of a negative
effect = 0.667), and the long-term interaction was $-0.020$ (95% CrI $-0.121$
to $0.081$; posterior probability of a negative effect = 0.649). Thus, both
posterior means point weakly in the previously reported direction, but the
intervals are centred close to zero and include effects of either sign.

## Predicted interaction

```{r}
#| label: interaction-figure
#| fig-cap: "Predicted EPP across temperature anomalies at representative absolute latitudes. Lines and ribbons are posterior medians and 95% credible intervals; points show observed brood-level proportions."
show_saved_figure("figures/interaction_temperature_anomaly_by_latitude.png",
                  "Interaction prediction figure will appear after model fitting.")
```

The prediction grid explicitly supplies `n_broods_sampled = 21`, the median in
the 441-record fitting frame. Posterior expected counts are divided by 21, so
the plotted response is a proportion rather than an expected brood count.

```{r}
#| label: interaction-prediction-scale
if (!is.na(prediction_scale_file)) kable(read_csv_safe(prediction_scale_file))
```

## Diagnostics

```{r}
#| label: interaction-diagnostics
if (!is.na(diagnostics_file)) kable(read_csv_safe(diagnostics_file), digits = 2)
```

Both fits mixed well: the maximum R-hat was 1.003, minimum bulk and tail ESS
were above 1,300, and neither model had divergent transitions or maximum-
treedepth hits.

## LOO comparison with main-effects models

Each interaction model is compared only with its matching main-effects-only counterpart. Pareto-`k` diagnostics are reported separately and are not replaced silently by another method.

```{r}
#| label: interaction-loo
if (!is.na(loo_file)) kable(read_csv_safe(loo_file), digits = 2)
if (!is.na(pareto_file)) kable(read_csv_safe(pareto_file), digits = 2)
```

The main-effects-only models had slightly higher expected predictive accuracy
than their interaction counterparts (ELPD differences of 1.92 for the short
anomaly and 1.51 for the long anomaly, with the main-effects model as the
reference). These LOO comparisons are explicitly treated as approximate:
several observations had Pareto $k>0.7$ and each interaction fit had one
observation with $k>1$. The posterior coefficient and its uncertainty therefore
remain the primary evidence for the confirmatory interaction tests.

### Observation-level Pareto-*k* audit

High Pareto-*k* values were inspected against latitude and study intensity,
rather than only counted. Across the interaction analysis, none of the
problematic observations belonged to the five best-studied populations and
their median population record count was one. They therefore are not an
artefact of a few intensively sampled European populations. Tropical records
were overrepresented: they comprise 8.4% of the full frame but 22--40% of the
high-*k* sets, depending on the matched model. Recurring observations included
*Ardea ibis* at absolute latitude 22.5 degrees (observed EPP proportion 0.613;
maximum *k* 1.206) and *Spheniscus humboldti* at 15.37 degrees (observed EPP
proportion 0). This identifies sparse, biologically unusual low-latitude
records as an important source of predictive difficulty.

```{r}
#| label: interaction-pareto-audit
if (!is.na(pareto_audit_summary_file)) {
  pareto_summary <- read_csv_safe(pareto_audit_summary_file) |>
    filter(analysis == "interaction") |>
    select(model, max_pareto_k, n_pareto_k_gt_0_7, n_pareto_k_gt_1,
           n_high_tropical, proportion_high_tropical,
           n_high_top5_population, median_population_records_high)
  kable(pareto_summary, digits = 3,
        caption = "Model-level summary of inspected Pareto-k observations.")
}
if (!is.na(pareto_audit_high_file)) {
  pareto_cases <- read_csv_safe(pareto_audit_high_file) |>
    filter(analysis == "interaction") |>
    arrange(desc(max_pareto_k)) |>
    select(species, family, order, abs_latitude, tropical,
           population_record_count, n_broods_sampled, observed_epp_rate,
           models_high, max_pareto_k)
  kable(pareto_cases, digits = 3,
        caption = "Unique observations with Pareto k above 0.7 in an interaction or matched main-effects model.")
}
```

Because these approximate LOO comparisons are not the evidential basis for the
interaction conclusion, their ELPD ranking is not carried into the manuscript
as a formal model-selection result. If an ELPD comparison is retained in the
submitted manuscript, exact `reloo` will be run for the flagged observations;
the posterior interaction coefficients and intervals themselves do not depend
on that future calculation.

## Fixed effects

```{r}
#| label: interaction-fixed-effects
if (!is.na(effects_file)) kable(read_csv_safe(effects_file), digits = 3)
```
