---
title: "Figures"
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(brms)
library(dplyr)
library(posterior)
library(ggplot2)
library(grid)
library(knitr)
```

# Aim

This page creates the figures used throughout the book.

The variation figures summarise biological structure in EPP, and the forest plot compares effects from the absolute climate, short-term anomaly, and long-term anomaly models. All continuous predictors were standardised before model fitting. Model effects are shown on the logit scale.

# Output Directory

```{r}
#| label: create-figure-directory

dir.create("figures", recursive = TRUE, showWarnings = FALSE)
```

# First-Month Versus Full-Season Variables

```{r}
#| label: figure-first-month-full-season-data

dat_model_phylo <- readRDS("data/processed/dat_model_phylo.rds")

make_first_full_data <- function(data, first_var, full_var, label) {
  keep <- complete.cases(data[, c(first_var, full_var)])
  test <- cor.test(
    data[[first_var]][keep],
    data[[full_var]][keep],
    method = "pearson"
  )

  data.frame(
    variable = label,
    first_month = data[[first_var]][keep],
    full_season = data[[full_var]][keep],
    r = unname(test$estimate),
    n = sum(keep)
  )
}

first_full_plot_data <- bind_rows(
  make_first_full_data(
    dat_model_phylo,
    "tmp_cru_C_first_month",
    "tmp_cru_C_full_season",
    "Temperature"
  ),
  make_first_full_data(
    dat_model_phylo,
    "pre_cru_mm_first_month",
    "pre_cru_mm_full_season",
    "Precipitation"
  ),
  make_first_full_data(
    dat_model_phylo,
    "dtr_cru_C_first_month",
    "dtr_cru_C_full_season",
    "DTR"
  ),
  make_first_full_data(
    dat_model_phylo,
    "ndvi_first_month_median_30km",
    "ndvi_full_season_median_30km",
    "NDVI"
  )
)

first_full_labels <- first_full_plot_data %>%
  group_by(variable) %>%
  summarise(
    label = paste0("r = ", round(first(r), 2), "; n = ", first(n)),
    x = min(first_month, na.rm = TRUE),
    y = max(full_season, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: figure-first-month-full-season-correlations
#| fig-width: 8
#| fig-height: 6

if (!exists("first_full_plot_data")) {
  if (!exists("dat_model_phylo")) {
    dat_model_phylo <- readRDS("data/processed/dat_model_phylo.rds")
  }

  make_first_full_data <- function(data, first_var, full_var, label) {
    keep <- complete.cases(data[, c(first_var, full_var)])
    test <- cor.test(
      data[[first_var]][keep],
      data[[full_var]][keep],
      method = "pearson"
    )

    data.frame(
      variable = label,
      first_month = data[[first_var]][keep],
      full_season = data[[full_var]][keep],
      r = unname(test$estimate),
      n = sum(keep)
    )
  }

  first_full_plot_data <- bind_rows(
    make_first_full_data(
      dat_model_phylo,
      "tmp_cru_C_first_month",
      "tmp_cru_C_full_season",
      "Temperature"
    ),
    make_first_full_data(
      dat_model_phylo,
      "pre_cru_mm_first_month",
      "pre_cru_mm_full_season",
      "Precipitation"
    ),
    make_first_full_data(
      dat_model_phylo,
      "dtr_cru_C_first_month",
      "dtr_cru_C_full_season",
      "DTR"
    ),
    make_first_full_data(
      dat_model_phylo,
      "ndvi_first_month_median_30km",
      "ndvi_full_season_median_30km",
      "NDVI"
    )
  )
}

if (!exists("first_full_labels")) {
  first_full_labels <- first_full_plot_data %>%
    group_by(variable) %>%
    summarise(
      label = paste0("r = ", round(first(r), 2), "; n = ", first(n)),
      x = min(first_month, na.rm = TRUE),
      y = max(full_season, na.rm = TRUE),
      .groups = "drop"
    )
}

fig_first_full_correlations <- ggplot(
  first_full_plot_data,
  aes(x = first_month, y = full_season)
) +
  geom_point(
    alpha = 0.35,
    size = 1.5,
    colour = "#6f7f62"
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linewidth = 0.7,
    colour = "#33412f"
  ) +
  geom_text(
    data = first_full_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 3.5,
    colour = "#24301f"
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    x = "First-month value",
    y = "Full-season value",
    title = "First-month environmental values track full-season conditions",
    subtitle = "Each panel compares the first breeding-season month with the corresponding full-season summary."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", colour = "#24301f"),
    plot.subtitle = element_text(colour = "#52624a"),
    strip.text = element_text(face = "bold", colour = "#24301f"),
    axis.title = element_text(colour = "#2f382c"),
    axis.text = element_text(colour = "#2f382c"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "#e3e7dc", linewidth = 0.35)
  )

fig_first_full_correlations

ggsave(
  "figures/first_month_full_season_correlations.png",
  fig_first_full_correlations,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(
  "figures/first_month_full_season_correlations.pdf",
  fig_first_full_correlations,
  width = 8,
  height = 6
)
```

# Variation Across Biological Levels

```{r}
#| label: figure-prepare-variation-data

weighted_mean <- function(x, w) {
  keep <- !is.na(x) & !is.na(w) & w > 0
  sum(x[keep] * w[keep]) / sum(w[keep])
}

dat_var <- dat_model_phylo %>%
  filter(
    !is.na(n_epp_broods),
    !is.na(n_broods_sampled),
    n_broods_sampled > 0,
    n_epp_broods <= n_broods_sampled,
    !is.na(species_phylo),
    !is.na(population_id)
  ) %>%
  mutate(
    epp_rate = n_epp_broods / n_broods_sampled,
    year_of_study_start = as.numeric(year_of_study_start),
    year_of_study_end = as.numeric(year_of_study_end),
    study_year = ifelse(
      !is.na(year_of_study_end),
      rowMeans(
        cbind(year_of_study_start, year_of_study_end),
        na.rm = TRUE
      ),
      year_of_study_start
    )
  )

species_epp <- dat_var %>%
  group_by(species_phylo) %>%
  summarise(
    species_mean_epp = weighted_mean(epp_rate, n_broods_sampled),
    n_populations = n_distinct(population_id),
    .groups = "drop"
  )

global_mean_epp <- weighted_mean(
  dat_var$epp_rate,
  dat_var$n_broods_sampled
)

population_epp <- dat_var %>%
  group_by(species_phylo, population_id) %>%
  summarise(
    population_mean_epp = weighted_mean(epp_rate, n_broods_sampled),
    n_years = n_distinct(study_year[!is.na(study_year)]),
    .groups = "drop"
  ) %>%
  left_join(
    species_epp %>%
      select(species_phylo, species_mean_epp, n_populations),
    by = "species_phylo"
  )

repeated_population_ids <- population_epp %>%
  filter(n_years >= 3) %>%
  pull(population_id)

year_epp <- dat_var %>%
  filter(population_id %in% repeated_population_ids) %>%
  left_join(
    population_epp %>%
      select(population_id, population_mean_epp),
    by = "population_id"
  ) %>%
  group_by(population_id) %>%
  summarise(
    abs_deviation = mean(
      abs(epp_rate - population_mean_epp),
      na.rm = TRUE
    ),
    .groups = "drop"
  )

variation_deviations <- bind_rows(
  species_epp %>%
    transmute(
      level = "Among species",
      unit = as.character(species_phylo),
      abs_deviation = abs(species_mean_epp - global_mean_epp)
    ),
  population_epp %>%
    filter(n_populations >= 3) %>%
    transmute(
      level = "Among populations within species",
      unit = as.character(population_id),
      abs_deviation = abs(population_mean_epp - species_mean_epp)
    ),
  year_epp %>%
    transmute(
      level = "Among years within population",
      unit = as.character(population_id),
      abs_deviation = abs_deviation
    )
)

variation_deviations$level <- factor(
  variation_deviations$level,
  levels = c(
    "Among species",
    "Among populations within species",
    "Among years within population"
  )
)
```

```{r}
#| label: figure-epp-variation-levels
#| fig-width: 8
#| fig-height: 5.5

fig_epp_variation_levels <- ggplot(
  variation_deviations,
  aes(x = level, y = abs_deviation)
) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    fill = "#dfe7d6",
    colour = "#33412f"
  ) +
  geom_jitter(
    width = 0.12,
    alpha = 0.35,
    size = 1.4,
    colour = "#6f7f62"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(100 * x), "%")
  ) +
  labs(
    x = NULL,
    y = "Absolute deviation in EPP",
    title = "EPP variation occurs across biological levels",
    subtitle = "Deviation is measured from the global mean, species mean, or population mean depending on level."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", colour = "#24301f"),
    plot.subtitle = element_text(colour = "#52624a"),
    axis.text.x = element_text(colour = "#2f382c"),
    axis.text.y = element_text(colour = "#2f382c"),
    axis.title.y = element_text(colour = "#2f382c"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(colour = "#e3e7dc", linewidth = 0.35)
  )

fig_epp_variation_levels

ggsave(
  "figures/epp_variation_across_levels.png",
  fig_epp_variation_levels,
  width = 8,
  height = 5.5,
  dpi = 300
)

ggsave(
  "figures/epp_variation_across_levels.pdf",
  fig_epp_variation_levels,
  width = 8,
  height = 5.5
)
```

# Family-Level Mean and Scale

```{r}
#| label: figure-family-mean-scale
#| fig-width: 8
#| fig-height: 5.8

if ("family" %in% names(dat_var) && any(!is.na(dat_var$family))) {
  species_family_epp <- dat_var %>%
    filter(
      !is.na(family),
      !is.na(species_phylo)
    ) %>%
    group_by(family, species_phylo) %>%
    summarise(
      species_mean_epp = weighted_mean(epp_rate, n_broods_sampled),
      total_broods = sum(n_broods_sampled, na.rm = TRUE),
      n_records = n(),
      .groups = "drop"
    )

  family_species_variation <- species_family_epp %>%
    group_by(family) %>%
    summarise(
      n_species = n_distinct(species_phylo),
      mean_species_epp = mean(species_mean_epp, na.rm = TRUE),
      sd_species_epp = sd(species_mean_epp, na.rm = TRUE),
      min_species_epp = min(species_mean_epp, na.rm = TRUE),
      max_species_epp = max(species_mean_epp, na.rm = TRUE),
      range_species_epp = max_species_epp - min_species_epp,
      .groups = "drop"
    ) %>%
    filter(n_species >= 3) %>%
    arrange(desc(sd_species_epp))

  fig_family_mean_scale <- ggplot(
    family_species_variation,
    aes(
      x = mean_species_epp,
      y = sd_species_epp,
      size = n_species
    )
  ) +
    geom_smooth(
      method = "lm",
      se = TRUE,
      linewidth = 0.7,
      colour = "#5f8c72",
      fill = "#dfe7d6"
    ) +
    geom_point(
      shape = 21,
      fill = "#d9b86c",
      colour = "#2f382c",
      alpha = 0.9,
      stroke = 0.35
    ) +
    geom_text(
      aes(label = family),
      size = 3,
      hjust = -0.08,
      vjust = 0.4,
      colour = "#2f382c",
      check_overlap = TRUE
    ) +
    scale_x_continuous(
      labels = function(x) paste0(round(100 * x), "%"),
      expand = expansion(mult = c(0.08, 0.24))
    ) +
    scale_y_continuous(
      labels = function(x) paste0(round(100 * x), "%"),
      expand = expansion(mult = c(0.08, 0.12))
    ) +
    scale_size_continuous(
      range = c(3, 8),
      name = "Species"
    ) +
    labs(
      x = "Family mean EPP",
      y = "Among-species EPP variation within family",
      title = "Family-level mean and scale of EPP",
      subtitle = "Families with higher mean EPP tended to show greater among-species divergence."
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", colour = "#24301f"),
      plot.subtitle = element_text(colour = "#52624a"),
      axis.title = element_text(colour = "#2f382c"),
      axis.text = element_text(colour = "#2f382c"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "#e3e7dc", linewidth = 0.35),
      legend.position = "bottom"
    )

  fig_family_mean_scale

  ggsave(
    "figures/family_mean_scale_epp.png",
    fig_family_mean_scale,
    width = 8,
    height = 5.8,
    dpi = 300
  )

  ggsave(
    "figures/family_mean_scale_epp.pdf",
    fig_family_mean_scale,
    width = 8,
    height = 5.8
  )
} else {
  cat(
    "Family-level mean-scale figure was not created because no family column is present in the current dataset.\n"
  )
}
```

# Load Fitted Models

```{r}
#| label: load-models

m_basic_pop <- readRDS("models/m_basic_pop.rds")
m_env_absolute <- readRDS("models/m_env_absolute.rds")
m_env_short <- readRDS("models/m_env_short.rds")
m_env_long <- readRDS("models/m_env_long.rds")
```

# Posterior Predictive Checks

```{r}
#| label: posterior-predictive-check-figures
#| fig-width: 8
#| fig-height: 5

pp_check_overall_basic_pop <- pp_check(
  m_basic_pop,
  type = "bars",
  ndraws = 100
)

pp_check_env_absolute <- pp_check(
  m_env_absolute,
  type = "bars",
  ndraws = 100
)

pp_check_env_short <- pp_check(
  m_env_short,
  type = "bars",
  ndraws = 100
)

pp_check_env_long <- pp_check(
  m_env_long,
  type = "bars",
  ndraws = 100
)

pp_check_overall_basic_pop
pp_check_env_absolute
pp_check_env_short
pp_check_env_long

ggsave(
  "figures/pp_check_overall_basic_pop.png",
  pp_check_overall_basic_pop,
  width = 8,
  height = 5,
  dpi = 300
)

ggsave(
  "figures/pp_check_overall_basic_pop.pdf",
  pp_check_overall_basic_pop,
  width = 8,
  height = 5
)

ggsave(
  "figures/pp_check_env_absolute.png",
  pp_check_env_absolute,
  width = 8,
  height = 5,
  dpi = 300
)

ggsave(
  "figures/pp_check_env_absolute.pdf",
  pp_check_env_absolute,
  width = 8,
  height = 5
)

ggsave(
  "figures/pp_check_env_short.png",
  pp_check_env_short,
  width = 8,
  height = 5,
  dpi = 300
)

ggsave(
  "figures/pp_check_env_short.pdf",
  pp_check_env_short,
  width = 8,
  height = 5
)

ggsave(
  "figures/pp_check_env_long.png",
  pp_check_env_long,
  width = 8,
  height = 5,
  dpi = 300
)

ggsave(
  "figures/pp_check_env_long.pdf",
  pp_check_env_long,
  width = 8,
  height = 5
)
```

# Extract Fixed Effects

```{r}
#| label: extract-effects

extract_fixed_effects <- function(model, model_label, term_labels) {
  draws <- as_draws_df(model)

  available_terms <- intersect(names(term_labels), names(draws))

  do.call(
    rbind,
    lapply(
      available_terms,
      function(term) {
        x <- draws[[term]]

        data.frame(
          model = model_label,
          term = term,
          predictor = term_labels[[term]],
          estimate = median(x),
          lower = quantile(x, 0.025),
          upper = quantile(x, 0.975),
          prob_positive = mean(x > 0),
          prob_negative = mean(x < 0),
          row.names = NULL
        )
      }
    )
  )
}

moderator_terms <- c(
  b_abs_latitude_z = "Absolute latitude",
  b_distance_to_coast_z = "Distance to coast",
  b_nest_boxyes = "Nest box",
  b_marker_typemicrosatellite = "Marker: microsatellite",
  b_marker_typeSNPs = "Marker: SNPs",
  b_marker_typeother = "Marker: other"
)

absolute_terms <- c(
  b_tmp_cru_C_first_month_z = "Temperature",
  b_pre_cru_mm_first_month_log_z = "Precipitation",
  b_dtr_cru_C_first_month_z = "DTR",
  b_ndvi_first_month_median_30km_z = "NDVI",
  moderator_terms
)

short_terms <- c(
  b_tmp_short_anom_5yr_C_first_month_z = "Temperature anomaly",
  b_pre_short_anom_5yr_mm_first_month_z = "Precipitation anomaly",
  b_dtr_cru_C_first_month_z = "DTR",
  b_ndvi_first_month_median_30km_z = "NDVI",
  moderator_terms
)

long_terms <- c(
  b_tmp_long_anom_1961_1989_C_first_month_z = "Temperature anomaly",
  b_pre_long_anom_1961_1989_mm_first_month_z = "Precipitation anomaly",
  b_dtr_cru_C_first_month_z = "DTR",
  b_ndvi_first_month_median_30km_z = "NDVI",
  moderator_terms
)

effects_list <- list(
  extract_fixed_effects(m_env_absolute, "Absolute climate", absolute_terms),
  extract_fixed_effects(m_env_short, "Short anomalies", short_terms),
  extract_fixed_effects(m_env_long, "Long anomalies", long_terms)
)

effects_env <- do.call(rbind, effects_list)

model_levels <- c(
  "Absolute climate",
  "Short anomalies",
  "Long anomalies"
)

effects_env$model <- factor(
  effects_env$model,
  levels = model_levels[model_levels %in% effects_env$model]
)

predictor_order <- c(
  "Temperature",
  "Temperature anomaly",
  "Precipitation",
  "Precipitation anomaly",
  "DTR",
  "NDVI",
  "Absolute latitude",
  "Distance to coast",
  "Nest box",
  "Marker: microsatellite",
  "Marker: SNPs",
  "Marker: other"
)

effects_env$predictor <- factor(
  effects_env$predictor,
  levels = rev(predictor_order)
)

effects_env$credible_direction <- ifelse(
  effects_env$lower > 0 | effects_env$upper < 0,
  "95% CrI excludes zero",
  "95% CrI overlaps zero"
)

kable(
  effects_env[
    ,
    c(
      "model",
      "predictor",
      "estimate",
      "lower",
      "upper",
      "prob_positive",
      "prob_negative"
    )
  ],
  digits = 3
)
```

# Forest Plot

```{r}
#| label: forest-plot
#| fig-width: 9
#| fig-height: 9

fig_environmental_effects <- ggplot(
  effects_env,
  aes(x = estimate, y = predictor)
) +
  geom_vline(
    xintercept = 0,
    linewidth = 0.4,
    linetype = "dashed",
    colour = "grey35"
  ) +
  geom_errorbarh(
    aes(xmin = lower, xmax = upper, colour = credible_direction),
    height = 0,
    linewidth = 0.7
  ) +
  geom_point(
    aes(fill = credible_direction),
    shape = 21,
    size = 2.8,
    colour = "black",
    stroke = 0.25
  ) +
  facet_wrap(~ model, ncol = 1, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "95% CrI excludes zero" = "#1b9e77",
      "95% CrI overlaps zero" = "grey45"
    )
  ) +
  scale_fill_manual(
    values = c(
      "95% CrI excludes zero" = "#1b9e77",
      "95% CrI overlaps zero" = "white"
    )
  ) +
  labs(
    x = "Posterior effect on EPP probability (logit scale)",
    y = NULL,
    colour = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", hjust = 0),
    legend.position = "bottom",
    axis.text.y = element_text(size = 10),
    panel.spacing.y = unit(1.1, "lines")
  )

fig_environmental_effects
```

# Save Figure

```{r}
#| label: save-figure

dir.create("figures", recursive = TRUE, showWarnings = FALSE)

ggsave(
  filename = "figures/main_environmental_effects_forest_plot.png",
  plot = fig_environmental_effects,
  width = 9,
  height = 9,
  dpi = 300
)

ggsave(
  filename = "figures/main_environmental_effects_forest_plot.pdf",
  plot = fig_environmental_effects,
  width = 9,
  height = 9
)
```
