---
title: "Comparison with Botero & Rubenstein (2012)"
---

## Overview

Botero and Rubenstein (2012) analysed the proportion of nests containing
extra-pair young, the same response definition used here, but aggregated the
data to one value per species. Their only retained climate association for
infidelity was a weak positive effect of within-year temperature variance
($\beta = 0.042$, $P = 0.041$) in a final model with $R^2 = 0.08$. This chapter
tests that predictor directly in the present population-resolved,
beta-binomial phylogenetic framework and then reproduces their species-level
PGLS workflow on our data. The comparison is framed as a test of
generalisability, not as a claim that the earlier result was strong.

::: {.callout-warning title="Scope of the comparison: context, not a formal replication"}
Only **44 of the 122 species** in B&R's infidelity dataset occur in the
present annual-data model set. This limited overlap is driven primarily by
our requirement for year-resolved population records, not by failed taxonomy:
all 122 B&R names were resolved to Clements v2025. In addition, adult
mortality and coloniality, which entered B&R's candidate model, are not
available consistently in our dataset. Our models instead retain the
pre-specified phylogenetic, population, marker-type, and nest-box controls.

Consequently, the analyses in this chapter provide **comparative context and
a test of generalisability**, not a formal replication of B&R's complete
analysis. The separate Table S1 check below validates the PGLS implementation
on their original response and predictors; it does not remove the differences
in species coverage, covariates, or phylogeny.
:::

```{r}
#| label: br-setup
#| echo: true

library(dplyr)
library(readr)
library(knitr)

model_dir <- "models/br_comparison_v409"
legacy_model_dir <- "models/br_comparison"

# Base R's reader is used here because vroom/readr can fail to open otherwise
# valid relative paths when the Windows project path contains non-ASCII text.
read_csv_safe <- function(path) {
  utils::read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}

show_saved_summary <- function(summary_file, label) {
  if (file.exists(summary_file)) {
    cat("<details><summary>", label, "</summary>\n\n```\n", sep = "")
    cat(readLines(summary_file), sep = "\n")
    cat("\n```\n</details>\n")
  } else {
    cat("Saved model summary is not available yet.")
  }
}

show_csv <- function(filename, digits = 3) {
  path <- file.path(model_dir, filename)
  if (file.exists(path)) {
    knitr::kable(read_csv_safe(path), digits = digits)
  } else {
    cat("Saved output is not available yet.")
  }
}

show_legacy_csv <- function(filename, digits = 3) {
  path <- file.path(legacy_model_dir, filename)
  if (file.exists(path)) {
    knitr::kable(read_csv_safe(path), digits = digits)
  } else {
    cat("Saved version-invariant output is not available yet.")
  }
}
```

## Climate reconstruction

Monthly CRU TS v4.09 temperature and precipitation were extracted at the
audited 0.5-degree locality cells. The locality-to-cell assignments were held
fixed while changing the CRU product version, so the v4.09--v4.10 comparison
does not conflate product version with spatial remapping. All populations use
the fixed 1961--1990 reference window, preventing the derived climate
descriptors from being confounded with study year. No climate value was
imputed. Sites without a reported breeding-season end month retain the two
within-year metrics but are excluded from analyses requiring among-year
breeding-season variance.

Breeding seasons that cross December/January are treated as one continuous
season: months from the reported start through December use reference year
$y$, whereas January-side months use $y+1$. Thus, a season beginning in 1990
may use calendar months from 1991. The v4.09 extraction consequently includes
the 1991--2000 archives solely to complete the 1961--1990 reference seasons.
All 23 unique cross-year locality--season combinations (63 population-year
records) retained 30 complete temperature and precipitation seasons.

For variable $x$ in year $y$, within-year variance is
$s_y^2 = \operatorname{Var}(x_{y,1}, \ldots, x_{y,12})$; the locality value is
the mean of $s_y^2$ across 1961--1990. Among-year variance is the variance,
across those 30 years, of the mean over the reported breeding-season months.
The four variables are:

| Variable | Definition |
|---|---|
| `br_tmp_withinyear_var` | Mean annual variance of the 12 monthly temperatures |
| `br_pre_withinyear_var` | Mean annual variance of the 12 monthly precipitation totals |
| `br_tmp_amongyear_var` | Variance among annual breeding-season mean temperatures |
| `br_pre_amongyear_var` | Variance among annual breeding-season mean precipitation totals |

Each predictor is standardised only after complete-case filtering of its own
model frame. DTR is not included in the comparison models: daily thermal
amplitude and B&R's seasonal monthly variance are different quantities.

```{r}
#| label: br-source-manifests
cru_manifest_path <- file.path(model_dir, "cru_ts_4.09_file_manifest.csv")
if (file.exists(cru_manifest_path)) {
  read_csv_safe(cru_manifest_path) %>%
    mutate(sha256 = paste0(substr(sha256, 1, 12), "...")) %>%
    kable(digits = 0)
}

table_manifest_path <- file.path(legacy_model_dir, "br_table_s1_source_manifest.csv")
if (file.exists(table_manifest_path)) {
  read_csv_safe(table_manifest_path) %>%
    select(
      article, doi, pmcid, source_file, source_sheet,
      source_species_rows, retrieved_date
    ) %>%
    kable(digits = 0)
}
```

```{r}
#| label: br-model-frames
show_csv("br_model_frame_checks.csv", digits = 0)
show_csv("br_scaling_parameters.csv", digits = 3)
show_csv("br_record_level_season_audit.csv", digits = 0)
show_csv("br_v409_vs_v410_record_comparison.csv", digits = 3)
```

B1 and B1b retain all 441 records. B2 contains 426 records because 15 records
lack a reported breeding-season end month; this is the only loss introduced
by its among-year seasonal predictors. The version comparison is presented
as an audit, not as a model-selection exercise. In particular, precipitation
among-year variance changes only for cross-year seasons, as expected from the
calendar-year correction.

## Relation to the existing environmental predictors

The correlation panel distinguishes B&R's seasonal temperature variance from
our raw temperature, DTR, absolute latitude, and thermal-geographic PC1.

```{r}
#| label: br-correlation-panel
panel_path <- file.path(model_dir, "br_correlation_panel.csv")
if (file.exists(panel_path)) {
  panel <- read_csv_safe(panel_path) %>%
    filter(variable_1 == "br_tmp_withinyear_var", variable_2 != variable_1)
  kable(panel, digits = 3)
}
```

PC1 is not a pure temperature-variability axis. Its largest relevant loadings
are positive for mean temperature and DTR and negative for absolute latitude;
distance to coast also contributes. The raw-predictor sensitivity model is
therefore required to decide whether its weak positive effect can be assigned
to any one measured variable.

```{r}
#| label: br-pc1-resolution
loadings_path <- "models/pca_environment_model_set/pca_loadings_environment_geography.csv"
sensitivity_path <- "models/sensitivity_models/fixed_effects_sensitivity_models.csv"
if (file.exists(loadings_path)) {
  read_csv_safe(loadings_path) %>%
    select(variable, PC1_thermal_geographic) %>%
    arrange(desc(abs(PC1_thermal_geographic))) %>%
    kable(digits = 3)
}
if (file.exists(sensitivity_path)) {
  read_csv_safe(sensitivity_path) %>%
    filter(
      component == "mu",
      predictor %in% c("Mean temperature", "DTR", "Absolute latitude", "Distance to coast")
    ) %>%
    select(model, predictor, estimate, lower, upper, prob_positive, prob_negative) %>%
    kable(digits = 3)
}
```

## Validation of the PGLS pipeline on B&R Table S1

Before applying the species-level PGLS workflow to our data, we tested it on
the 122 original infidelity rows in B&R Table S1. The source workbook is
accepted only if its SHA-256 checksum and row count match the recorded source
manifest. Following the reported final infidelity model, the response is the
arcsine-square-root transformation of the proportion of broods with EPP;
adult mortality and within-year temperature variance are standardised; and a
PGLS with Pagel's lambda is fitted. This checksum-verified validation uses
B&R's published Table S1 values and is therefore invariant to our CRU product
version; its existing validated fit is retained rather than needlessly
refitted.

```{r}
#| label: br-pgls-validation-code
#| eval: false
m_br_table_s1_validation_pgls <- phylolm::phylolm(
  epp_arcsine_sqrt ~
    adult_mortality_z + within_year_temperature_variance_z,
  data = dat_br_table_s1,
  phy = phy_br_table_s1,
  model = "lambda"
)
```

```{r}
#| label: br-pgls-validation-results
show_legacy_csv("br_pgls_pipeline_validation.csv", digits = 4)
```

```{r}
#| label: br-pgls-validation-summary
#| results: asis
show_saved_summary(
  file.path(legacy_model_dir, "summary_m_br_table_s1_validation_pgls.txt"),
  "Show summary for the Table S1 validation PGLS"
)
```

The reproduced temperature-seasonality coefficient is required
programmatically to round to B&R's published value of **0.042**; a mismatch
stops the pipeline. The original RAxML tree is available in the article
package only as a plotted PDF rather than a machine-readable tree with branch
lengths. We therefore use the current dated Clements tree pruned to the 122
Table S1 species. The coefficient is the primary implementation check;
differences in the reproduced $t$, $P$, and lambda are retained in the table
as transparent secondary consequences of using a different phylogeny.

## Bayesian comparison models

The priors are reproduced literally from the main fitting script and compared
programmatically with its parsed `priors_mu` assignment before fitting. A
difference stops the script. All three models otherwise use the main model's
beta-binomial family, sampler settings, population random intercept,
non-phylogenetic species intercept, phylogenetic covariance, nest-box effect,
and marker-type effect. Coloniality and adult mortality were unavailable and
were not replaced by new proxies.

### B1: retained B&R predictor

```{r}
#| label: br-b1-code
#| eval: false
m_br_tmp_withinyear <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      br_tmp_withinyear_var_z + nest_box + marker_type +
      (1 | population_id) + (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_cov)),
    phi ~ 1
  ),
  data = dat_b1,
  data2 = list(phylo_cov = phylo_b1),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4, cores = 4, iter = 10000, warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 2411,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: br-b1-summary
#| results: asis
show_saved_summary(
  file.path(model_dir, "summary_m_br_tmp_withinyear.txt"),
  "Show summary for B1"
)
```

### B1b: retained B&R predictor adjusted for latitude

B1 is deliberately left without latitude because it is the direct,
head-to-head test of B&R's retained seasonality predictor. B1b is a separate
confounding test: it adds standardised absolute latitude to the otherwise
identical model. This distinction matters because within-year temperature
variance and absolute latitude are appreciably associated in these data
($r = 0.400$), precisely the concern raised for this literature by
[Brouwer and Griffith (2019)](https://doi.org/10.1111/mec.15259).

```{r}
#| label: br-b1b-code
#| eval: false
m_br_tmp_withinyear_latitude <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      br_tmp_withinyear_var_z + abs_latitude_z +
      nest_box + marker_type +
      (1 | population_id) + (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_cov)),
    phi ~ 1
  ),
  data = dat_b1,
  data2 = list(phylo_cov = phylo_b1),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4, cores = 4, iter = 10000, warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 2412,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: br-b1b-summary
#| results: asis
show_saved_summary(
  file.path(model_dir, "summary_m_br_tmp_withinyear_latitude.txt"),
  "Show summary for B1b"
)
```

### B2: full variance set

```{r}
#| label: br-b2-code
#| eval: false
m_br_full_climate <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      br_tmp_withinyear_var_z + br_pre_withinyear_var_z +
      br_tmp_amongyear_var_z + br_pre_amongyear_var_z +
      nest_box + marker_type +
      (1 | population_id) + (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_cov)),
    phi ~ 1
  ),
  data = dat_b2,
  data2 = list(phylo_cov = phylo_b2),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu,
  chains = 4, cores = 4, iter = 10000, warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 2413,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: br-b2-summary
#| results: asis
show_saved_summary(
  file.path(model_dir, "summary_m_br_full_climate.txt"),
  "Show summary for B2"
)
```

### Bayesian results and diagnostics

All four B2 climate coefficients are reported; no backward elimination is
used. Predictor VIFs and posterior coefficient correlations are retained so
that collinearity is reported rather than resolved by selecting a preferred
coefficient.

```{r}
#| label: br-bayesian-results
show_csv("br_b1_b1b_seasonality_comparison.csv", digits = 3)
show_csv("fixed_effects_br_models.csv", digits = 3)
show_csv("br_fixed_effects_v409_vs_v410.csv", digits = 3)
show_csv("diagnostics_br_models.csv", digits = 2)
show_csv("br_predictor_vif_b2.csv", digits = 2)
show_csv("br_posterior_coefficient_correlations_b2.csv", digits = 2)
```

B1 provides no support for the positive B&R association: the standardised
within-year temperature-variance coefficient is approximately $-0.05$
(95% CrI $[-0.23, 0.13]$). In B1b it remains null at approximately $-0.02$
($[-0.21, 0.17]$), while absolute latitude is also unsupported at
approximately $-0.11$ ($[-0.32, 0.10]$). Thus, adjustment for latitude does
not uncover an independent seasonality effect, and neither predictor explains
the other's null association. Exact posterior summaries and diagnostics are
reported in the tables above.

In B2, neither temperature coefficient is supported: within-year temperature
variance is $-0.085$ (95% CrI $[-0.296, 0.128]$) and among-year seasonal
temperature variance is $0.026$ ($[-0.162, 0.213]$). Within-year
precipitation variance is negative at $-0.221$ ($[-0.445, -0.001]$,
probability of direction $=0.976$), whereas among-year seasonal precipitation
variance is positive but uncertain at $0.190$ ($[-0.014, 0.394]$,
probability of direction $=0.966$). The former interval only narrowly excludes
zero and its v4.09 estimate differs from the v4.10 estimate by $-0.001$; the
threshold crossing is therefore not a material product-version effect. With
four correlated climate predictors and no backward elimination, it is treated
as a weak secondary signal rather than a selected independent effect.

## C1: Species-level PGLS on our data

To separate a methodological difference from a sampling difference, the
current observations are aggregated to species means, EPP is transformed as
$\arcsin(\sqrt{p})$, the climate predictors are standardised, and a PGLS with
Pagel's $\lambda$ is fitted. The model contains within-year temperature and
precipitation variance and uses the same Clements-matched phylogeny. This is a
supplementary decomposition of the discrepancy, not a preferred primary model.

```{r}
#| label: br-pgls-results
show_csv("coefficients_br_species_pgls.csv", digits = 4)
show_csv("summary_metrics_br_species_pgls.csv", digits = 4)
```

```{r}
#| label: br-pgls-summary
#| results: asis
show_saved_summary(
  file.path(model_dir, "summary_m_br_species_pgls.txt"),
  "Show summary for the species-level PGLS"
)
```

On the v4.09 reconstruction, neither species-level coefficient differs from
zero: within-year temperature variance is $-0.0035$ (SE $=0.0282$,
$P=0.902$) and within-year precipitation variance is $-0.0054$
(SE $=0.0255$, $P=0.833$; 110 species). Thus, applying B&R's species-level
aggregation and response transformation to the present data does not recover
their positive temperature-seasonality association.

## Species overlap with the original study

Species names from the 122-row infidelity sheet in B&R Table S1 were matched
to the local Clements v2025 checklist. Exact scientific-name matches were used
first, followed by exact English-name matches. Non-exact assignments were
accepted only when the relevant Clements split or infraspecific-group row was
verified automatically. The source spelling, Clements name, matching method,
and overlap indicator are all retained in the crosswalk.

```{r}
#| label: br-species-overlap-summary
show_legacy_csv("br_species_overlap_summary.csv", digits = 0)
```

```{r}
#| label: br-species-overlap-crosswalk
crosswalk_path <- file.path(legacy_model_dir, "br_species_overlap_crosswalk.csv")
if (file.exists(crosswalk_path)) {
  read_csv_safe(crosswalk_path) %>%
    select(
      br_scientific_name_original,
      br_common_name_original,
      clements_name,
      match_method,
      match_status,
      in_current_data
    ) %>%
    kable()
}
```

## Interpretation rule

The head-to-head conclusion is based on B1; B1b must not replace it. B1b asks
whether the seasonality coefficient contains information independent of
absolute latitude. If the coefficient is attenuated or loses posterior support
after adding latitude, while latitude itself remains unsupported in the main
latitude sensitivity analysis and in B1b, the B&R association is statistically
indistinguishable from a null latitudinal gradient. That is a separate result,
not evidence that latitude explains EPP.

If B1's 95% credible interval for within-year temperature variance includes
zero and the species-level PGLS is also null, the discrepancy is attributed
primarily to data and sampling rather than to species averaging or the arcsine
transformation. B&R's among-year
breeding-season variance was already absent from their final infidelity model;
our year-specific anomaly results therefore extend, rather than contradict,
that null. Any manuscript statement should retain the modest magnitude and
borderline support of the original result.

Source: [Botero & Rubenstein (2012)](https://doi.org/10.1371/journal.pone.0032311).
Confounding rationale: [Brouwer & Griffith (2019)](https://doi.org/10.1111/mec.15259).
