---
title: "Preregistered zero-inflated beta-binomial models"
---

```{r}
#| label: zi-setup
#| include: false
library(dplyr)
library(knitr)

model_dir <- "models/zero_inflated_models"
read_result <- function(name) utils::read.csv(file.path(model_dir, name), check.names = FALSE)
result_files <- c(
  "zero_inflation_parameter.csv", "all_fixed_effects_zero_inflated.csv",
  "fixed_effects_standard_vs_zero_inflated.csv",
  "loo_standard_vs_zero_inflated_reloo.csv",
  "reloo_diagnostics_zero_inflated_comparisons.csv",
  "reloo_vs_moment_match_comparison_change.csv",
  "diagnostics_zero_inflated_models.csv", "posterior_predictive_zero_counts.csv"
)
models_ready <- all(file.exists(file.path(model_dir, result_files)))
if (models_ready) {
  zi_parameter <- read_result("zero_inflation_parameter.csv")
  coefficients <- read_result("all_fixed_effects_zero_inflated.csv")
  comparison <- read_result("fixed_effects_standard_vs_zero_inflated.csv")
  loo_comparison <- read_result("loo_standard_vs_zero_inflated_reloo.csv")
  reloo_diagnostics <- read_result("reloo_diagnostics_zero_inflated_comparisons.csv")
  reloo_change <- read_result("reloo_vs_moment_match_comparison_change.csv")
  diagnostics <- read_result("diagnostics_zero_inflated_models.csv")
  zero_ppc <- read_result("posterior_predictive_zero_counts.csv")

  zi_reloo <- loo_comparison |>
    filter(version == "zero_inflated") |>
    mutate(model_kind = factor(model_kind, levels = c("pca", "short", "long"))) |>
    arrange(model_kind)
  zi_reloo_change <- reloo_change |>
    filter(version == "zero_inflated") |>
    mutate(model_kind = factor(model_kind, levels = c("pca", "short", "long"))) |>
    arrange(model_kind)
  if (nrow(zi_reloo) != 3L || any(zi_reloo$elpd_diff >= 0)) {
    stop("The exact-reloo result changed the written zero-inflation interpretation.")
  }
}
marker_zeros <- read_result("zero_records_by_marker_type.csv")
prior_audit <- read_result("priors_mu_automatic_verification.csv")
pareto_audit_summary_file <- file.path("models/pareto_k_audit", "pareto_k_model_summary.csv")
pareto_audit_high_file <- file.path("models/pareto_k_audit", "pareto_k_unique_high_observations.csv")
```

## Why fit a zero-inflated model when the zero PPC already passes?

This analysis implements the preregistered primary family without changing any of the existing main models. The diagnostic result preceding the refit is important: the standard beta-binomial PCA model observed 48 zero-EPP records but predicted a mean of 53.7 (95% posterior-predictive interval 41--66; posterior predictive probability of a count no larger than observed = 0.21). Thus the standard model already reproduces the zeros and, if anything, slightly overpredicts them. The zero-inflated fits below are retained because they were preregistered, not because the data show unexplained excess zeros.

Zeros also occurred mainly in small samples: the median sample was 11.5 broods for zero records and 23 for non-zero records, and no zero occurred above 50 sampled broods. This pattern is consistent with beta-binomial sampling variation. The raw marker contrast points in the registered direction but is not a controlled marker effect: DNA fingerprinting had 18.2% zero records and microsatellites 8.4%. Marker type is associated with study era and taxonomic composition, so these percentages are descriptive only.

```{r}
#| label: zi-marker-table
kable(marker_zeros, digits = 2, caption = "Observed zero records by marker type (descriptive).")
```

## Model specification and reproducibility safeguards

The PCA, short-anomaly, and long-anomaly mean models use `zero_inflated_beta_binomial(link = "logit", link_phi = "log", link_zi = "logit")`. Every other element is held fixed: the exact saved 441-row model frame, predictors, phylogenetic covariance, three random intercepts, four chains, 10,000 iterations (5,000 warm-up), `adapt_delta = 0.99`, and maximum tree depth 15. New seeds are 1601, 1603, and 1605.

The original `priors_mu` block was reproduced literally and automatically compared with the parsed assignment in the unchanged main-model script. This is an executable equality check rather than a manual assertion.

```{r}
#| label: zi-prior-audit
kable(prior_audit, caption = "Automatic prior-reproduction check.")
```

The additional zero-inflation intercept received a Normal(-2, 1.5) prior on the logit scale. Its prior median is 0.119 on the probability scale, while its width allows probabilities close to zero as well as appreciable inflation. Because a logit parameter maps continuously to the open interval (0, 1), its credible interval cannot literally include zero. Evidence against a needed structural-zero component must instead come from posterior concentration near zero, the zero-count PPC, and out-of-sample predictive comparison.

## Estimated structural-zero probabilities

```{r}
#| label: zi-parameter-table
kable(
  if (models_ready) zi_parameter else data.frame(status = "Model fitting is queued; no estimate is reported yet."),
  digits = 3,
  caption = "Posterior structural-zero probability in each preregistered model."
)
```

All three posterior structural-zero probabilities were approximately 0.0056,
with 95% CrIs from about 0.0008 to 0.016. In each model, roughly 87% of the
posterior mass lay below 0.01 and all of it lay below 0.05. This is
concentration very near zero, not evidence for a substantial structural-zero
process.

## All population-level coefficients

All coefficients are reported; no backward elimination was performed. Continuous coefficients are on their fitted standardized scales. Factor coefficients are contrasts against the reference level used in the unchanged main model frame.

```{r}
#| label: zi-all-coefficients
kable(
  if (models_ready) coefficients else data.frame(status = "Model fitting is queued; no coefficients are reported yet."),
  digits = 3,
  caption = "Population-level coefficients from the zero-inflated models (posterior median, 95% credible interval, and probability of direction)."
)
```

## Did substantive estimates change?

The table pairs each zero-inflated coefficient with the corresponding standard beta-binomial coefficient. `delta` is zero-inflated minus standard on the logit scale. A change is flagged as material only if the absolute posterior-median shift exceeds 0.1 and the 95% interval for the shift excludes zero; the full continuous comparison is retained so this reporting threshold does not replace interpretation.

```{r}
#| label: zi-side-by-side
kable(
  if (models_ready) comparison else data.frame(status = "Model fitting is queued; no comparison is reported yet."),
  digits = 3,
  caption = "Standard versus zero-inflated fixed effects."
)
```

No fixed-effect coefficient met the prespecified material-change criterion.

## Predictive comparison

LOO `r if (models_ready) "was" else "will be"` computed separately for each zero-inflated model and compared only with its standard counterpart fitted to the identical 441 observations. Initial PSIS-LOO estimates used moment matching. For the manuscript comparison, within each of the six fitted models, every observation with Pareto $k > 0.7$ was then replaced by an exact leave-one-out refit (`reloo`). This required `r if (models_ready) sum(reloo_diagnostics$n_exact_refits) else "pending"` refits in total: `r if (models_ready) sum(reloo_diagnostics$n_exact_refits[reloo_diagnostics$version == "standard"]) else "pending"` for the standard models and `r if (models_ready) sum(reloo_diagnostics$n_exact_refits[reloo_diagnostics$version == "zero_inflated"]) else "pending"` for the zero-inflated models. Each refit retained the original priors, formula, random effects, and sampler settings.

```{r}
#| label: zi-loo
kable(
  if (models_ready) {
    loo_comparison |>
      select(model_kind, version, elpd_diff, se_diff, elpd_loo, se_elpd_loo,
             p_loo, looic)
  } else {
    data.frame(status = "Exact reloo is queued; LOO is not available yet.")
  },
  digits = 2,
  caption = "Exact-reloo comparison: zero-inflated versus standard beta-binomial."
)
```

The standard beta-binomial was favoured in every matched comparison: the
zero-inflated fits had exact-reloo ELPD differences of
`r if (models_ready) sprintf("%.2f", zi_reloo$elpd_diff[zi_reloo$model_kind == "pca"]) else "pending"`
(PCA),
`r if (models_ready) sprintf("%.2f", zi_reloo$elpd_diff[zi_reloo$model_kind == "short"]) else "pending"`
(short anomaly), and
`r if (models_ready) sprintf("%.2f", zi_reloo$elpd_diff[zi_reloo$model_kind == "long"]) else "pending"`
(long anomaly). Together with the passing zero-count PPC and the near-zero
structural-zero posterior, this shows that the preregistered zero-inflation
component is unnecessary for these data. Relative to moment matching, exact
refitting made the zero-inflated-minus-standard ELPD difference more negative
by `r if (models_ready) sprintf("%.2f", -zi_reloo_change$reloo_minus_moment_match_elpd_diff[zi_reloo_change$model_kind == "pca"]) else "pending"`,
`r if (models_ready) sprintf("%.2f", -zi_reloo_change$reloo_minus_moment_match_elpd_diff[zi_reloo_change$model_kind == "short"]) else "pending"`, and
`r if (models_ready) sprintf("%.2f", -zi_reloo_change$reloo_minus_moment_match_elpd_diff[zi_reloo_change$model_kind == "long"]) else "pending"`
ELPD units for the PCA, short-anomaly, and long-anomaly pairs, respectively;
thus `reloo` strengthened rather than reversed the original conclusion.

Before exact refitting, each zero-inflated model had 6--8 observations above
0.7; the PCA and long-anomaly models each had one above 1. None came from the
five best-studied populations, while tropical records were again more frequent
among the flagged cases than in the full frame. These observations motivated
`reloo`; after their pointwise contributions were replaced by exact
leave-one-out estimates, no unresolved Pareto-$k$ value above 0.7 remained in
any of the six LOO objects used in the comparison.

```{r}
#| label: zi-pareto-audit
if (file.exists(pareto_audit_summary_file)) {
  zi_pareto_summary <- utils::read.csv(pareto_audit_summary_file, check.names = FALSE) |>
    filter(analysis == "zero_inflated") |>
    select(model, max_pareto_k, n_pareto_k_gt_0_7, n_pareto_k_gt_1,
           n_high_tropical, proportion_high_tropical,
           n_high_top5_population, median_population_records_high)
  kable(zi_pareto_summary, digits = 3,
        caption = "Observation-level Pareto-k audit for zero-inflated models.")
}
```

```{r}
#| label: zi-reloo-diagnostics
if (models_ready) {
  reloo_diagnostics |>
    select(model_kind, version, n_exact_refits, max_pareto_k_before,
           n_pareto_k_gt_1_before, n_pareto_k_gt_0_7_after,
           reloo_minus_psis_elpd, elapsed_minutes) |>
    kable(digits = 3,
          caption = "Exact-reloo coverage and change from moment-matched PSIS-LOO.")
}
```

```{r}
#| label: zi-zero-ppc
kable(if (models_ready) zero_ppc else data.frame(status = "Model fitting is queued; posterior predictions are not available yet."), digits = 2, caption = "Observed and posterior-predictive zero counts.")
```

## Sampler diagnostics

```{r}
#| label: zi-diagnostics
kable(if (models_ready) diagnostics else data.frame(status = "Model fitting is queued; sampler diagnostics are not available yet."), digits = 3, caption = "Sampler and posterior diagnostics.")
```

All three fits converged well (maximum R-hat 1.003, minimum ESS above 1,300,
no divergences, no maximum-treedepth hits, and no reported fit warnings).

The raw fits, summaries, coefficient comparison, zero-count checks, original moment-matched LOO objects, exact-reloo objects, pointwise replacement audits, prior audit, and session information are retained under `models/zero_inflated_models/` and `logs/` without overwriting the standard models.
