---
title: "Exploratory location-scale models"
description: "Exploratory models allowing predictors to affect the beta-binomial precision parameter."
---

## Overview

This chapter presents the exploratory location-scale versions of the three main environmental model groups. These models keep the same mean component as the corresponding main environmental models, but also allow selected environmental predictors and methodological controls to enter the beta-binomial precision component.

In `brms::beta_binomial(link_phi = "log")`, positive coefficients in the precision component indicate higher precision and lower residual overdispersion, whereas negative coefficients indicate lower precision and higher residual overdispersion. These models are used as diagnostic and sensitivity checks rather than as the primary presentation set.

```{r}
#| label: location-scale-setup
#| echo: true

library(dplyr)
library(readr)
library(tibble)
library(knitr)

read_csv_safe <- function(path) {
  tibble::as_tibble(utils::read.csv(
    path, check.names = FALSE, stringsAsFactors = FALSE
  ))
}

model_dir <- "models/main_environmental_models"

first_existing <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) {
    return(NA_character_)
  }
  existing[1]
}

summary_files <- list(
  pca_mu_phi = first_existing(c(file.path(model_dir, "summary_m_main_pca_mu_phi.txt"))),
  short_mu_phi = first_existing(c(file.path(model_dir, "summary_m_main_short_mu_phi.txt"))),
  long_mu_phi = first_existing(c(file.path(model_dir, "summary_m_main_long_mu_phi.txt")))
)

show_saved_summary <- function(summary_file, label) {
  if (!is.na(summary_file) && file.exists(summary_file)) {
    cat("<details><summary>", label, "</summary>\n\n```\n", sep = "")
    cat(readLines(summary_file), sep = "\n")
    cat("\n```\n</details>\n")
  } else {
    cat("Saved model summary is not available yet.")
  }
}
```

## Model set

| Hypothesis group | Location-scale model |
|----|----|
| Environmental PCA | Mean: `PC1_thermal_geographic + PC2_wet_productive + nest_box + marker_type`; precision: `phi ~ PC1_thermal_geographic + PC2_wet_productive + nest_box + marker_type` |
| Short anomaly | Mean: `tmp_short_anom_5yr_C_first_month_z + pre_short_anom_5yr_mm_first_month_z + abs_latitude_z + distance_to_coast_z + nest_box + marker_type`; precision: `phi ~ tmp_short_anom_5yr_C_first_month_z + pre_short_anom_5yr_mm_first_month_z + nest_box + marker_type` |
| Long anomaly | Mean: `tmp_long_anom_1961_1989_C_first_month_z + pre_long_anom_1961_1989_mm_first_month_z + abs_latitude_z + distance_to_coast_z + nest_box + marker_type`; precision: `phi ~ tmp_long_anom_1961_1989_C_first_month_z + pre_long_anom_1961_1989_mm_first_month_z + nest_box + marker_type` |
## Environmental PCA location-scale model

```{r}
#| label: main-pca-mu-phi-code
#| eval: false

m_main_pca_mu_phi <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      PC1_thermal_geographic +
      PC2_wet_productive +
      nest_box +
      marker_type +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat_fit)),

    phi ~
      PC1_thermal_geographic +
      PC2_wet_productive +
      nest_box +
      marker_type
  ),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1402,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-pca-mu-phi-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$pca_mu_phi, "Show summary for the Environmental PCA location-scale model")
```

## Short-anomaly location-scale model

```{r}
#| label: main-short-mu-phi-code
#| eval: false

m_main_short_mu_phi <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      tmp_short_anom_5yr_C_first_month_z +
      pre_short_anom_5yr_mm_first_month_z +
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat_fit)),

    phi ~
      tmp_short_anom_5yr_C_first_month_z +
      pre_short_anom_5yr_mm_first_month_z +
      nest_box +
      marker_type
  ),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1404,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-short-mu-phi-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$short_mu_phi, "Show summary for the short-anomaly location-scale model")
```

## Long-anomaly location-scale model

```{r}
#| label: main-long-mu-phi-code
#| eval: false

m_main_long_mu_phi <- brms::brm(
  brms::bf(
    n_epp_broods | trials(n_broods_sampled) ~
      tmp_long_anom_1961_1989_C_first_month_z +
      pre_long_anom_1961_1989_mm_first_month_z +
      abs_latitude_z +
      distance_to_coast_z +
      nest_box +
      marker_type +
      (1 | population_id) +
      (1 | species_nonphylo) +
      (1 | gr(species_phylo, cov = phylo_mat_fit)),

    phi ~
      tmp_long_anom_1961_1989_C_first_month_z +
      pre_long_anom_1961_1989_mm_first_month_z +
      nest_box +
      marker_type
  ),
  data = dat_fit,
  data2 = list(phylo_mat_fit = phylo_mat_fit),
  family = brms::beta_binomial(link = "logit", link_phi = "log"),
  prior = priors_mu_phi,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 1406,
  save_pars = brms::save_pars(all = TRUE)
)
```

```{r}
#| label: main-long-mu-phi-summary
#| echo: true
#| results: asis

show_saved_summary(summary_files$long_mu_phi, "Show summary for the long-anomaly location-scale model")
```

## Location-scale model diagnostics

The location-scale models were fitted to evaluate whether environmental predictors and methodological controls also explained variation in the beta-binomial precision parameter. In `brms::beta_binomial(link_phi = "log")`, positive coefficients in the precision component indicate higher precision and lower residual overdispersion, whereas negative coefficients indicate lower precision and higher residual overdispersion.

The fitted location-scale models did not show sampler pathologies: no divergent transitions or maximum-treedepth hits were detected. However, the scale-component coefficients were generally estimated with wide credible intervals that overlapped zero. These results therefore provide limited support for interpreting predictor-specific changes in residual overdispersion, and the location-scale models are treated as exploratory rather than as the primary presentation set.

```{r}
#| label: main-location-scale-sampler-diagnostics
#| echo: true

location_scale_diagnostics_file <- file.path(
  model_dir,
  "location_scale_sampler_diagnostics_from_rds.csv"
)

if (file.exists(location_scale_diagnostics_file)) {
  read_csv_safe(location_scale_diagnostics_file) %>%
    mutate(
      model = recode(
        model,
        m_main_pca_mu_phi = "Environmental PCA location-scale",
        m_main_short_mu_phi = "Short-anomaly location-scale",
        m_main_long_mu_phi = "Long-anomaly location-scale"
      )
    ) %>%
    rename(
      `Model` = model,
      `Divergent transitions` = divergent_transitions,
      `Maximum treedepth hits` = treedepth_hits
    ) %>%
    kable()
} else {
  cat("Location-scale sampler diagnostics are not available yet.")
}
```

```{r}
#| label: main-location-scale-phi-effects
#| echo: true

location_scale_phi_file <- file.path(
  model_dir,
  "location_scale_phi_effects_from_rds.csv"
)

phi_term_labels <- c(
  b_phi_PC1_thermal_geographic = "PC1 thermal-geographic",
  b_phi_PC2_wet_productive = "PC2 wet-productivity",
  b_phi_tmp_short_anom_5yr_C_first_month_z = "Short temperature anomaly",
  b_phi_pre_short_anom_5yr_mm_first_month_z = "Short precipitation anomaly",
  b_phi_tmp_long_anom_1961_1989_C_first_month_z = "Long temperature anomaly",
  b_phi_pre_long_anom_1961_1989_mm_first_month_z = "Long precipitation anomaly",
  b_phi_nest_boxyes = "Nest box",
  b_phi_marker_typemicrosatellite = "Marker: microsatellite",
  b_phi_marker_typeSNPs = "Marker: SNPs",
  b_phi_marker_typeother = "Marker: other"
)

if (file.exists(location_scale_phi_file)) {
  read_csv_safe(location_scale_phi_file) %>%
    filter(term != "b_phi_Intercept") %>%
    mutate(
      model = recode(
        model,
        m_main_pca_mu_phi = "Environmental PCA location-scale",
        m_main_short_mu_phi = "Short-anomaly location-scale",
        m_main_long_mu_phi = "Long-anomaly location-scale"
      ),
      term = recode(term, !!!phi_term_labels),
      across(c(estimate, lower, upper, prob_positive, prob_negative), ~ round(.x, 3))
    ) %>%
    select(
      `Model` = model,
      `Precision predictor` = term,
      Estimate = estimate,
      `Lower 95% CrI` = lower,
      `Upper 95% CrI` = upper,
      `Pr(positive)` = prob_positive,
      `Pr(negative)` = prob_negative
    ) %>%
    kable()
} else {
  cat("Location-scale precision-effect summaries are not available yet.")
}
```

